---
title: "Migration note: argument renaming in 19.15.0"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migration note: argument renaming in 19.15.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# make deprecation warnings fire deterministically for this document
options(lifecycle_verbosity = "warning")
```

```{r setup, message = FALSE}
library(SciDataReportR)
```

SciDataReportR 19.15.0 standardized argument names across the whole
package to snake_case. **Function names did not change** (still
PascalCase), **argument positions did not change**, and **every old
argument name still works** - it just emits one deprecation warning per
call pointing to the new name.

## What was renamed

The same concept used to go by several names in different functions.
Each cluster now has one name everywhere:

| Concept | Old names (varied by function) | New name |
|---|---|---|
| Primary data frame | `Data`, `DataFrame`, `df`, `Dataframe`, `DatatoRevalue` | `data` |
| Variable selection | `Variables`, `variables` | `variables` |
| Covariates | `Covariates`, `Covars`, `covars`, `covariates` | `covariates` |
| Merge keys | `Keys`, `MergeBy`, `key` | `keys` (`safe_merge()` keeps `by` to mirror `dplyr::left_join()`) |
| Participant ID column | `ID`, `id`, `id_col` | `id_var` |
| Grouping variable | `GroupVariable`, `GroupVar`, `Group`, `CompVariable` | `group_var` |
| Codebook | `Codebook`, `CB`, `VariableCodebook`, `VarTypes` | `codebook` |
| Digits | `numdecimals` | `digits` |
| Digits (multi-knob tables) | `ValueDigits`, `pDigits`, `EffectSizeDigits`, `TooltipDigits` | `value_digits`, `p_digits`, `effect_size_digits`, `tooltip_digits` |
| Interactive output | `Interactive`, `MakeInteractive`, `make_interactive` | `interactive` (each function keeps its original default) |
| Predictors (directional analyses) | `xVars`, `PredictorVars`, `predictorVars`, `predictorVar` | `predictor_vars` / `predictor_var` |
| Outcomes (directional analyses) | `yVars`, `yVar`, `OutcomeVars`, `outcomeVars`, `outcomeVar` | `outcome_vars` / `outcome_var` |
| Symmetric variable set | `xVars`/`yVars` in `PlotDirectionalHeatmaps()` | `variables` |

Also new in 19.15.0: `CreatePCAObject()`'s duplicate `imputeMethod`
argument is deprecated in favor of `ImputeMethod`, and the heatmap/matrix
family gained an `fdr_scope` argument (see the
*FDR correction scope* vignette).

## What you will see

Old code keeps working. The first thing you'll notice is a one-line
warning per deprecated argument, telling you the replacement:

```{r, warning = TRUE}
old_style <- MakeDataDictionary(DataFrame = mtcars, numdecimals = 2)
```

The new-style call is silent and returns the identical result:

```{r}
new_style <- MakeDataDictionary(mtcars, digits = 2)
identical(old_style, new_style)
```

Positional calls never changed, because every argument kept its position
(the deprecated names were appended at the end of each signature):

```{r}
identical(MakeDataDictionary(mtcars, 2), new_style)
```

```{r}
# a directional example: xVars/yVars became predictor_vars/outcome_vars
res <- PlotCorrelationsHeatmap(
  mtcars,
  predictor_vars = c("hp", "wt"),
  outcome_vars   = "mpg"
)
round(res$Unadjusted$p, 4)
```

## Migrating

There is no urgency - deprecated arguments will keep working throughout
the 19.x series. To migrate a script, rename arguments per the table
above; nothing else changes. If you want deprecation warnings to be
errors while you migrate (to catch every old usage), set:

```{r, eval = FALSE}
options(lifecycle_verbosity = "error")
```
