---
title: "Evaluating candidate biomarkers"
author: "Raha Dastgheyb"
title-block-banner: true
date: "last-modified"
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    smooth-scroll: true
    theme: zephyr
execute:
  echo: false
  message: false
  warning: false
  cache: false
vignette: >
  %\VignetteIndexEntry{Evaluating candidate biomarkers}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

This vignette uses the bundled `SampleData` to illustrate a realistic
candidate-biomarker workflow: separating participants with `Impaired` versus
`Control` diagnosis. It is a teaching dataset, not a validated diagnostic study.
The examples show apparent performance in this dataset and should not be used as
clinical cutoffs or estimates expected to transport unchanged to another cohort.

Performance is multidimensional. AUC describes discrimination, Brier score
describes probability accuracy, calibration asks whether predicted and observed
risk agree, and delta AUC asks whether a biomarker adds information beyond the
baseline covariate model.

## Load packages

```{r}
#| label: LoadPackages
library(SciDataReportR)
library(dplyr)
library(plotly)
```

## Load and revalue the bundled data

```{r}
#| label: LoadData
data("SampleData")
data("SampleVariableTypes")

df_Raw <- SampleData
VariableTypes <- SampleVariableTypes

df_Revalued <- RevalueData(df_Raw, VariableTypes)$RevaluedData
```

::: callout-note
The data file being used is `SampleData` from SciDataReportR.

The codebook being used is `SampleVariableTypes` from SciDataReportR.
:::

`Diagnosis` has two observed levels here: `Control` and `Impaired`. Every
binary example below explicitly treats `Impaired` as the positive outcome, so
the scientific interpretation cannot silently flip with factor order.

## Amyloid beta 42 as a candidate biomarker

Amyloid beta 42 is a useful first candidate in this teaching dataset. We adjust
for age because age can itself discriminate diagnostic groups and may account
for part of the biomarker-outcome association.

```{r}
#| label: AmyloidBeta42
Amyloid42Result <- EvaluateBiomarkerPerformance(
  data = df_Revalued,
  outcome_var = "Diagnosis",
  biomarker_var = "Ab_42",
  covariates = "age",
  PositiveLevel = "Impaired",
  codebook = VariableTypes
)

Amyloid42Result$PerformanceTable
```

Read the three rows together: `Biomarker` is amyloid beta 42 alone,
`Covariates` is the age-only baseline, and `Adjusted` combines both. The AUC
and Brier score describe different aspects of performance. A coefficient or AUC
that looks promising does not alone establish clinical value.

```{r}
#| label: AmyloidBeta42Plots
Amyloid42Result$ThresholdTable
Amyloid42Result$Plots$ROC
Amyloid42Result$Plots$Calibration
Amyloid42Result$Plots$CalibrationGrouped
Amyloid42Result$Plots$Probability
```

The raw threshold is expressed in amyloid beta 42 units; model-probability
thresholds are on a different scale. Sensitivity and specificity depend on the
selected threshold. PPV and NPV also depend on the prevalence in this dataset,
so they should not be assumed to apply to another setting. The probability curve
uses marginal standardization: each participant retains their observed age while
amyloid beta 42 is varied, then predicted probabilities are averaged.

## Compare several candidate proteins

In practice, the question is often comparative: does amyloid beta 42 outperform
tau, phosphorylated tau, serum amyloid P, or a weak comparator such as AXL?

```{r}
#| label: BiomarkerScreen
vars_Biomarkers <- c("Ab_42", "tau", "p_tau", "Serum_Amyloid_P", "AXL")

BiomarkerScreen <- ScreenBiomarkerPerformance(
  data = df_Revalued,
  outcome_vars = "Diagnosis",
  biomarker_vars = vars_Biomarkers,
  covariates = "age",
  PositiveLevel = "Impaired",
  codebook = VariableTypes
)

BiomarkerScreen$PerformanceTable
```

```{r}
#| label: ComparativePerformancePlots
BiomarkerScreen$Plots$AUC
BiomarkerScreen$Plots$IncrementalAUC
BiomarkerScreen$Plots$Heatmap
```

These screen-level plots provide a compact ranking. Each row has its own `N`
because analyses use the complete observations available for that biomarker,
outcome, and age. Compare AUC confidence intervals and missingness before
over-interpreting small differences.

```{r}
#| label: BiomarkerPanels
BiomarkerScreen$Plots$BiomarkerPanels
BiomarkerScreen$Plots$ROCFacets
```

The distribution panels are visual QC: they show the raw biomarker values in
each diagnostic group and annotate the adjusted AUC, confidence interval, and
pair-specific N. The ROC facets show adjusted discrimination for the same
candidate set. Neither plot turns a screen into external validation.

```{r}
#| label: InteractiveHeatmap
ggplotly(BiomarkerScreen$Plots$Heatmap, tooltip = "text")

BiomarkerScreen$Plots$Heatmap %>%
  add_biomarker_values()
```

## Multicategory categorical predictor: genotype

`Genotype` is a multicategory categorical predictor. It is evaluated through
the fitted model and predicted probabilities at each genotype level. A single raw
diagnostic cutoff would not make scientific sense, so it is intentionally absent.

```{r}
#| label: GenotypePredictor
GenotypeResult <- EvaluateBiomarkerPerformance(
  data = df_Revalued,
  outcome_var = "Diagnosis",
  biomarker_var = "Genotype",
  covariates = "age",
  PositiveLevel = "Impaired",
  codebook = VariableTypes
)

GenotypeResult$RawBiomarkerThreshold
GenotypeResult$PerformanceTable
GenotypeResult$Plots$Probability
```

## Continuous-outcome appendix

`SampleData` does not contain a credible continuous clinical endpoint. This
small simulated appendix exists only to demonstrate the continuous-outcome API;
the main workflow above remains entirely based on bundled sample data.

```{r}
#| label: ContinuousOutcomeData
set.seed(2026)
df_ContinuousExample <- tibble(
  Age = rnorm(150, 60, 10),
  Biomarker = rnorm(150)
) %>%
  dplyr::mutate(
    SymptomScore = 0.35 * .data$Biomarker + 0.03 * .data$Age + rnorm(150)
  )

ContinuousResult <- EvaluateBiomarkerPerformance(
  data = df_ContinuousExample,
  outcome_var = "SymptomScore",
  biomarker_var = "Biomarker",
  covariates = "Age",
  OutcomeType = "continuous"
)

ContinuousResult$PerformanceTable
ContinuousResult$Plots$PredictedObserved
```

Continuous outcomes use R-squared, adjusted R-squared, incremental R-squared,
RMSE, and MAE. ROC and classification metrics are intentionally not calculated.

## Internal validation

```{r}
#| label: CrossValidation
#| eval: false
ValidatedResult <- EvaluateBiomarkerPerformance(
  data = df_Revalued,
  outcome_var = "Diagnosis",
  biomarker_var = "Ab_42",
  covariates = "age",
  PositiveLevel = "Impaired",
  Validation = "cross_validation",
  CVFolds = 5,
  codebook = VariableTypes
)

ValidatedResult$Validation
ValidatedResult$ValidationPredictions
```

Apparent performance is evaluated in the same data used to fit the model and can
be optimistic. Cross-validation uses out-of-fold predictions; bootstrap output
is labelled optimism-corrected. Internal validation is still not external
validation in a new population.

## Reproducibility

```{r}
#| label: Reproducibility
print(sessionInfo())
# save.image()
```
