---
title: "Merge QC workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Merge QC workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
has_survival <- requireNamespace("survival", quietly = TRUE)
has_widgets <- requireNamespace("reactable", quietly = TRUE) &&
  requireNamespace("htmltools", quietly = TRUE)
knitr::opts_chunk$set(eval = has_survival)
```

Merging datasets is where many analysis errors are born: silently duplicated
rows, IDs that never match, and `.x` / `.y` column pairs that nobody resolves.
The damage is quiet — a join that duplicates rows inflates your sample without
error, and the analysis that follows is wrong in a way no downstream check will
catch.

SciDataReportR wraps the merge-and-audit cycle into a small workflow:

- `safe_merge()` performs a left join (exact-key or closest-time) and
  immediately audits the result with `ValidateMerge()`.
- `merge_detail()` prints plain-text diagnostic tables for one merge.
- `PlotMergeValidation()` draws the same diagnostics as figures.
- `ExploreMergeValidation()` renders an interactive dashboard for one merge.
- `merge_summary_table()` stacks the one-row logs from many merges into a
  pipeline-level rollup.

## The workflow

Every merge goes through the same four steps. The point of the workflow is
that step 2 is never skipped, and that the answer is written down rather than
eyeballed.

**1. State what you expect before merging.** `expected_relationship` is the
argument that turns an assumption into a check. If you believe each left-hand
row matches at most one right-hand row, say `"many-to-one"`; the audit then
compares that against what actually happened instead of leaving you to notice.

**2. Merge and audit in one call.** `safe_merge()` joins and validates
together, so an unaudited merge is not something you can produce by forgetting
a line.

**3. Read the log, then drill into whatever it flags.** The one-row `$log` is
the verdict; `merge_detail()`, `PlotMergeValidation()`, and
`ExploreMergeValidation()` are three views of the same underlying
`$validation` object, for three different situations.

**4. Roll the logs up at the end.** `merge_summary_table(..., flagged_only =
TRUE)` over every merge in the pipeline is the gate: an empty table means
every merge passed.

### What to check, and what it means

| Check | The question it answers | When to worry |
|---|---|---|
| **Row count** (`RowsBefore` / `RowsAfter`) | Did the join duplicate rows? | Any increase you did not intend. This is the silent killer: no error, larger sample, wrong analysis |
| **Relationship** (`DetectedRelationship`) | Was the join one-to-one, many-to-one, one-to-many? | It disagrees with `expected_relationship` |
| **Match rate** (`MatchRate`) | What fraction of left-hand keys found a match? | Below what the study design implies — a low rate usually means a key-format problem, not genuinely absent data |
| **Key types** (`KeyHarmonization`) | Were the keys the same type on both sides? | IDs stored as integer in one file and double or character in another; coercion happened, so check it was the coercion you wanted |
| **Duplicate keys** (`DuplicateKeyGroups`) | Does the key uniquely identify a row after merging? | Depends entirely on intent: fatal for one-row-per-participant data, expected for longitudinal data |
| **Unresolved duplicate variables** (`UnresolvedDupVars`) | Did the merge leave `.x` / `.y` column pairs? | Always. Every one is a column whose correct value nobody has decided on |
| **Value conflicts** | Where a variable came from both sources, do they agree? | Disagreement means the two sources disagree about a fact; find out which is right before analyzing either |

### Choosing a review surface

The three review functions read the same `$validation` object and differ only
in how they present it:

| Function | Output | Reach for it when |
|---|---|---|
| `merge_detail()` | Static `knitr::kable()` tables | The merge is part of a rendered report or log, and the record needs to survive in the document |
| `PlotMergeValidation()` | ggplot figures | Coverage and conflict *patterns* matter more than exact counts, or the result is going in a slide |
| `ExploreMergeValidation()` | Interactive HTML dashboard | You are working through a problem merge yourself and want to expand, sort, and filter rather than re-run code |

For a merge that passes, the `$log` alone is enough and none of the three is
needed.

```{r setup, message = FALSE}
library(SciDataReportR)
library(survival)
```

## Example 1: a real longitudinal merge (survival::pbc + pbcseq)

The `survival` package ships two related datasets from the Mayo Clinic primary
biliary cholangitis trial:

- `pbc`: baseline data, one row per patient (418 patients, including 106 who
  did not participate in the randomized trial),
- `pbcseq`: longitudinal follow-up labs, multiple rows per patient (only the
  312 trial participants).

Merging them by patient `id` is a completely ordinary task — and it surfaces
real coverage gaps and duplicate-key situations that were not staged for this
vignette.

```{r}
pbc_baseline <- pbc[, c("id", "age", "sex", "trt", "stage")]
pbc_labs <- pbcseq[, c("id", "day", "bili", "albumin", "platelet")]

m_pbc <- safe_merge(
  pbc_baseline,
  pbc_labs,
  by = "id",
  name = "pbc baseline + follow-up labs"
)
```

`safe_merge()` returns four elements and prints nothing on its own. The
one-row `log` is the pipeline-friendly view:

```{r}
m_pbc$log
```

The `summary` element is a ready-made `knitr::kable()`:

```{r}
m_pbc$summary
```

The audit correctly reports this merge as `FAIL` — not because the code is
wrong, but because the result needs a decision from you before analysis:

- **Coverage**: 106 patients exist only in the baseline data (the
  non-trial patients who have no follow-up labs).
- **Duplicate keys**: `pbcseq` has many rows per patient, so `id` alone does
  not uniquely identify a row after the merge. If you intended a one-row-per-
  patient dataset, this is a genuine error; if you intended a longitudinal
  dataset, you now have written confirmation of what happened.

### Drilling in with merge_detail()

`merge_detail()` prints plain kable tables for the checks, the unmatched keys
on each side, overlapping variables, and suspicious conflicts, skipping any
empty section. In R Markdown, use a chunk with `results = "asis"`:

```{r, results = "asis"}
merge_detail(m_pbc, TopN = 5)
```

### A closest-time variant

Suppose instead you want a one-row-per-patient dataset containing the lab
draw closest to enrollment. `safe_merge(method = "closest_time")` calls
`Merge_ByClosestTime()` under the hood, matching on `id` exactly and picking
the `pbc_labs` row whose `day` is nearest to each baseline row's time
(enrollment is day 0 here).

```{r}
pbc_enroll <- pbc_baseline
pbc_enroll$enroll_day <- 0

m_closest <- safe_merge(
  pbc_enroll,
  pbc_labs,
  by = "id",
  name = "closest lab to enrollment",
  method = "closest_time",
  time_var_before = "enroll_day",
  time_var_add = "day"
)

m_closest$log
```

Note the caveat: `ValidateMerge()` audits duplicate keys using `by` alone
(here `id`), not the time variables. In longitudinal settings, repeated keys
in the lab data are legitimate repeated visits, so a duplicate-key `FAIL` from
a closest-time merge may reflect expected repetition rather than a merge
error. The validation output is deliberately left as-is — read it with that
caveat in mind (here, the merged data itself has one row per patient, which
you can confirm from `RowsBefore` and `RowsAfter`).

## Example 2: a synthetic merge that fails every check

To demonstrate the remaining check types — key-type mismatches, unresolved
`.x` / `.y` pairs, and value conflicts — here is a small synthetic pair of
data frames constructed to go wrong in all the classic ways: a duplicated
key, an ID stored as integer on one side and double on the other, IDs
missing from each side, and a `sex` column present in both sources with
disagreeing values.

```{r}
demographics <- data.frame(
  id = 1:6, # integer
  sex = c("F", "M", "F", "F", "M", "F"),
  age = c(54, 61, 47, 66, 58, 50)
)

device_data <- data.frame(
  id = c(1, 2, 2, 5, 7), # double, with id 2 duplicated
  sex = c("F", "F", "M", "M", "F"), # disagrees with demographics for id 2/5
  score = c(0.82, 0.75, 0.71, 0.64, 0.90)
)

m_synth <- safe_merge(
  demographics,
  device_data,
  by = "id",
  name = "demographics + device data"
)

m_synth$log
```

```{r}
m_synth$summary
```

Every check type in `ValidateMerge()` now has something to say: the duplicate
key and the unresolved `sex.x` / `sex.y` pair are integrity blockers
(`FAIL`), while key-type coercion, coverage gaps, row inflation, overlapping
variables, and value conflicts appear as warnings.

```{r, results = "asis"}
merge_detail(m_synth)
```

### The same merge as figures

`PlotMergeValidation()` renders the identical diagnostics as ggplot figures.
`Plot = "All"` returns them as a named list, so a report can include only the
ones that matter for the merge at hand.

```{r}
plots_synth <- PlotMergeValidation(
  m_synth$validation,
  Plot = "All",
  interactive = FALSE
)
names(plots_synth)
```

The checks figure is the plotted form of the same traffic-light table that
`merge_detail()` prints first:

```{r, fig.width = 7, fig.height = 4}
plots_synth$Checks
```

Coverage shows where the keys came from — matched on both sides, or present in
only one. A pattern here is often more diagnostic than the counts: keys missing
from one side in a block usually points at a formatting or provenance problem
rather than genuinely absent records.

```{r, fig.width = 7, fig.height = 4}
plots_synth$Coverage
```

### Interactive review with ExploreMergeValidation()

For interactive QC sessions, pass the `validation` element to
`ExploreMergeValidation()`. This is the same content again, but explorable:
the default `Detail = "Compact"` keeps the checks table front and center, with
the coverage and conflict explorers as click-to-expand accordion sections
labeled with their item counts. Use `Detail = "Full"` to render them expanded.

Use this one while you are still working out *what went wrong*; use
`merge_detail()` once you know, and need the answer written into a report.

```{r, eval = has_survival && has_widgets}
ExploreMergeValidation(
  m_synth$validation,
  Title = "Demographics + device data",
  Detail = "Compact"
)
```

## Pipeline rollup with merge_summary_table()

After a sequence of merges, stack the logs into a single table. With
`flagged_only = TRUE`, only merges whose worst check status is not `PASS`
remain — an end-of-pipeline QC gate:

```{r}
merge_summary_table(
  list(m_pbc$log, m_closest$log, m_synth$log),
  flagged_only = TRUE
)
```

An empty table here would mean every merge in the pipeline passed cleanly;
any row that appears is a merge that still needs review before analysis.

## Stating the expectation up front

The single highest-value habit in this workflow is declaring
`expected_relationship` before the merge runs. Without it, the audit can
report what happened but not whether that was what you wanted:

```{r}
m_declared <- safe_merge(
  pbc_baseline,
  pbc_labs,
  by = "id",
  name = "pbc, declared as many-to-one",
  expected_relationship = "many-to-one"
)

m_declared$log[, c(
  "ExpectedRelationship", "DetectedRelationship",
  "RelationshipMatchesExpected"
)]
```

Here the declaration is wrong and the audit says so: `pbcseq` has many rows per
patient, so the join is one-to-many, not many-to-one. That mismatch is exactly
the signal worth having — it is the difference between a longitudinal dataset
you meant to build and a baseline dataset you have accidentally duplicated.

## Summary

1. Declare `expected_relationship` before merging.
2. Merge with `safe_merge()`, which audits in the same call.
3. Read `$log`. If it is `PASS`, move on.
4. If not, drill in with `merge_detail()` for a report,
   `PlotMergeValidation()` for the pattern, or `ExploreMergeValidation()` to
   investigate interactively.
5. Gate the pipeline with
   `merge_summary_table(logs, flagged_only = TRUE)`.

A `FAIL` is not necessarily a bug. It means the merge produced something that
needs a decision from you — and the value of the workflow is that the decision
is recorded rather than assumed.
