---
title: "Working with ordinal variables"
author: "Raha Dastgheyb"
title-block-banner: true
date: "last-modified"
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    smooth-scroll: true
    theme: zephyr
execute:
  echo: true
  message: false
  warning: false
  cache: false
vignette: >
  %\VignetteIndexEntry{Working with ordinal variables}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Why use more than one ordinal perspective?

Ordinal variables have a meaningful order, but their adjacent categories are
not automatically equally spaced. A symptom scale with *None*, *Mild*,
*Moderate*, and *Severe* can therefore be useful as a categorical outcome when
we care about the distribution across levels, or as a continuous score when a
one-step increase is a clinically reasonable working scale.

Neither view is universally correct. Looking at both is often useful during
EDA: the categorical result shows whether a pattern is concentrated in one
level, while the continuous result is a compact summary of an ordered trend.
The appropriate confirmatory analysis should still be chosen from the study
question, measurement properties, and prespecified analysis plan.

## Load packages

```{r}
#| label: LoadPackages
#| echo: false
library(SciDataReportR)
library(dplyr)
```

## Create a labelled ordinal example

```{r}
#| label: LoadData
#| echo: false
set.seed(20260802)

df_Raw <- data.frame(
  Group = factor(rep(c("Control", "Disease"), each = 45)),
  SymptomSeverity = c(
    sample(0:3, 45, replace = TRUE, prob = c(0.45, 0.30, 0.18, 0.07)),
    sample(0:3, 45, replace = TRUE, prob = c(0.12, 0.28, 0.35, 0.25))
  ),
  Age = round(stats::rnorm(90, mean = 48, sd = 12))
)

VariableTypes <- data.frame(
  Variable = c("Group", "SymptomSeverity", "Age"),
  Label = c("Participant group", "Symptom Severity", "Age (years)"),
  Type = c("Categorical", "Ordinal", "Continuous"),
  Recode = c("No", "Yes", "No"),
  Code = c(NA, "0=None;1=Mild;2=Moderate;3=Severe", NA)
)

df_Revalued <- RevalueData(df_Raw, VariableTypes)$RevaluedData
```

::: callout-note
This vignette constructs a small example dataset in the document. The
`VariableTypes` object above is the codebook used by `RevalueData()`.
:::

`RevalueData()` keeps `SymptomSeverity` as an ordered factor for categorical
analysis and records its numeric codebook scores (`0` through `3`) for a
continuous analysis. If a codebook has no complete numeric mapping, the ordered
level ranks are used instead.

## Comparison tables

`TreatOrdinalAs` is the common control for tables:

| Value | Table behavior |
|---|---|
| `"Categorical"` | Show ordinal levels and use a categorical test. |
| `"Continuous"` | Use the preserved numeric score and continuous summary/test. |
| `"Both"` | Show separate categorical and continuous entries. |
| `"Exclude"` | Omit ordinal variables from automatic selection. |

```{r}
#| label: CategoricalComparisonTable
tbl_Categorical <- MakeComparisonTable(
  data = df_Revalued,
  group_var = "Group",
  variables = "SymptomSeverity",
  TreatOrdinalAs = "Categorical"
)

tbl_Categorical
```

```{r}
#| label: ContinuousComparisonTable
tbl_Continuous <- MakeComparisonTable(
  data = df_Revalued,
  group_var = "Group",
  variables = "SymptomSeverity",
  TreatOrdinalAs = "Continuous"
)

tbl_Continuous
```

```{r}
#| label: BothComparisonTable
tbl_Both <- MakeComparisonTable(
  data = df_Revalued,
  group_var = "Group",
  variables = "SymptomSeverity",
  TreatOrdinalAs = "Both"
)

tbl_Both
```

With `TreatOrdinalAs = "Both"`, the original data are not changed. The table
uses temporary analysis columns and displays them as **Symptom Severity
(categorical)** and **Symptom Severity (continuous)**.
With `Relabel = FALSE`, the same entries use the raw variable name instead:
`SymptomSeverity (categorical)` and `SymptomSeverity (continuous)`.

## Plotting options

Plots intentionally allow only representations that make sense for the plot.

| Function | Ordinal parameter | Valid values |
|---|---|---|
| `PlotCategoricalDistributions()` | `TreatOrdinalAs` | `"Categorical"`, `"Exclude"` |
| `PlotContinuousDistributions()` | `TreatOrdinalAs` | `"Continuous"`, `"Exclude"` |
| `PlotCorrelationsHeatmap()` | `TreatOrdinalAs` | `"Continuous"`, `"Exclude"` |
| `PlotMiningMatrix()` | `TreatOrdinalAs` | all four values |

`"Both"` is rejected by single-representation plots because it would make the
analytic question ambiguous. Use separate categorical and continuous plots
when you want to inspect both perspectives.

```{r}
#| label: CategoricalOrdinalDistribution
PlotCategoricalDistributions(
  data = df_Revalued,
  variables = "SymptomSeverity",
  TreatOrdinalAs = "Categorical"
)
```

```{r}
#| label: ContinuousOrdinalDistribution
PlotContinuousDistributions(
  data = df_Revalued,
  variables = "SymptomSeverity",
  TreatOrdinalAs = "Continuous"
)
```

## PlotMiningMatrix() with both representations

`PlotMiningMatrix()` is the exception: it can examine mixed variable types, so
`TreatOrdinalAs = "Both"` adds two temporary analysis variables for each
ordinal field. It evaluates each representation wherever a valid pairwise test
is available; a particular representation will not appear when no compatible
variable pair exists.

```{r}
#| label: OrdinalMiningMatrix
#| column: screen
vars_Mining <- c("Group", "SymptomSeverity", "Age")

MiningObj <- PlotMiningMatrix(
  data = df_Revalued,
  outcome_vars = vars_Mining,
  TreatOrdinalAs = "Both",
  Relabel = TRUE
)

MiningObj$Unadjusted$plot
```

The plot labels use the same suffixes as the table: **Symptom Severity
(categorical)** and **Symptom Severity (continuous)**.
This makes the statistical interpretation visible rather than hiding it behind
an internal duplicate-column name. Set `Relabel = FALSE` to use raw variable
names with those same suffixes.

## Reproducibility

```{r}
#| label: Reproducibility
#| echo: false
#| output: false
# save.image()
print(sessionInfo())
```
