---
title: "Forest Plots from Univariate Regression Tables"
description: "Visualize estimates from MakeUnivariateRegressionTable() with PlotForestFromTable()."
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Forest Plots from Univariate Regression Tables}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

`PlotForestFromTable()` visualizes the estimates and confidence intervals from a `MakeUnivariateRegressionTable()` object. It plots from the tidy `Results` dataframe that `MakeUnivariateRegressionTable()` returns, and it also accepts that dataframe directly, so you can filter or reorder results before plotting.

The default layout is:

* facets = outcomes
* rows = predictors or predictor levels
* x-axis = regression estimate

Use `Flip = TRUE` when it is easier to read outcomes as rows and predictors as facets.

::: {.callout-note}
`PlotForestFromTable()` was previously named `plotForestFromTable()`. The old name still works as a backwards-compatible synonym, so existing scripts continue to run.
:::

## Load packages

```{r}
#| label: LoadPackages
library(SciDataReportR)
library(dplyr)
```

## Load example data

The examples use `SampleData` and `SampleVariableTypes`, then apply labels and recoding with `RevalueData()`.

```{r}
#| label: LoadData
data("SampleData")
data("SampleVariableTypes")

RevaluedObj <- RevalueData(
  SampleData,
  SampleVariableTypes
)

df_Revalued <- RevaluedObj$RevaluedData
```

## Many outcomes and predictors

This is a common screening pattern: several outcomes tested against several predictors.

```{r}
#| label: FitOutcomeScreen
vars_Outcomes <- c(
  "Calbindin",
  "Ferritin",
  "MMP7",
  "Sortilin"
)

vars_Predictors <- c(
  "Diagnosis",
  "age"
)

Reg_Obj_Outcomes <- MakeUnivariateRegressionTable(
  data = df_Revalued,
  outcome_vars = vars_Outcomes,
  predictor_vars = vars_Predictors,
  Standardize = TRUE
)
```

## Default forest plot

The default plot keeps outcomes as facets and terms as rows.

```{r}
#| label: PlotDefaultForest
PlotForestFromTable(Reg_Obj_Outcomes)
```

Black points indicate `p < 0.05`; grey points are not significant. The dashed vertical line is the null value for linear estimates.

## Plot from the Results dataframe

The plot is drawn from `Reg_Obj_Outcomes$Results`, and you can pass that dataframe (or any subset of it) directly. This makes it easy to plot only the rows you care about.

```{r}
#| label: PlotFilteredResults
Reg_Obj_Outcomes$Results %>%
  filter(Predictor == "Diagnosis") %>%
  PlotForestFromTable()
```

## Flip outcomes and predictors

When there are many outcomes and a small predictor set, `Flip = TRUE` is often easier to read.

```{r}
#| label: PlotFlippedForest
PlotForestFromTable(
  Reg_Obj_Outcomes,
  Flip = TRUE
)
```

With `Flip = TRUE`, predictors or predictor levels become facets, and outcomes become rows.

## Many predictors, one outcome

For one outcome and many predictors, the default layout is usually already compact.

```{r}
#| label: FitPredictorScreen
Reg_Obj_Predictors <- MakeUnivariateRegressionTable(
  data = df_Revalued,
  outcome_vars = "Sortilin",
  predictor_vars = c("Diagnosis", "age", "Calbindin", "Ferritin", "MMP7"),
  Standardize = TRUE
)
```

```{r}
#| label: PlotManyPredictors
PlotForestFromTable(Reg_Obj_Predictors)
```

## Labels

Forest plot facets and rows use labels inherited from `SampleVariableTypes` through `RevalueData()`. If labels are missing, the plot falls back to the variable names.

```{r}
#| label: InspectPlotLabels
Reg_Obj_Outcomes$Metadata$Outcomes
```

Categorical and dichotomous predictors are shown with the modeled level, for example `Diagnosis : Impaired`.

## Logistic models

`MakeUnivariateRegressionTable()` uses logistic regression automatically for two-level outcomes and reports odds ratios by default.

```{r}
#| label: FitLogisticScreen
Reg_Obj_Logistic <- MakeUnivariateRegressionTable(
  data = df_Revalued,
  outcome_vars = "Diagnosis",
  predictor_vars = c("age", "Calbindin", "Ferritin", "MMP7"),
  Standardize = TRUE
)

Reg_Obj_Logistic$FormattedTable
```

```{r}
#| label: InspectLogisticMetadata
Reg_Obj_Logistic$Metadata$Outcomes
```

```{r}
#| label: PlotLogisticForest
PlotForestFromTable(Reg_Obj_Logistic)
```

For logistic models, check the event level in `Metadata$Outcomes` before interpreting odds ratios. Mixed linear and logistic forest plots can be visually convenient, but they combine different effect scales.

## Reproducibility

```{r}
#| label: Reproducibility
# save.image("PlotForestFromTable_workspace.RData")
print(sessionInfo())
```
