---
title: "Clustering and clinical phenotyping"
author: "Raha Dastgheyb"
date: "last-modified"
bibliography: ../inst/REFERENCES.bib
format:
  html:
    minimal: true
    embed-resources: true
    toc: true
    toc-location: left
    smooth-scroll: true
execute:
  echo: false
  message: false
  warning: false
  cache: false
vignette: >
  %\VignetteIndexEntry{Clustering and clinical phenotyping}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

<style>
table {
  width: 100%;
  table-layout: fixed;
  font-size: 0.68rem;
  line-height: 1.2;
}
table th, table td {
  padding: 0.35rem 0.45rem;
  overflow-wrap: anywhere;
}
</style>

# Purpose

Clinical phenotyping uses prespecified features to discover structure in a
training cohort, then places later participants in that **fixed** phenotype
space. Clustering is exploratory: fit indices, internal stability, and clean
projection do not establish that a phenotype is biologically real. Clinical
interpretation and independent-cohort validation remain essential.

Every pipeline in this package uses the same train-once/project-many contract:

- preprocessing and model parameters are learned from training data only;
- `.scidr_rowid` preserves row alignment and full-length results;
- incomplete cases remain in the output with `NA` assignments;
- bootstrap stability refits the **entire pipeline**, including scaling and
  PCA/MCA when present;
- fit, membership uncertainty or distance, and projection support are
  reviewed before phenotypes are named.

`method = "exploratory"` (or the shorter alias `"explore"`) compares candidate
settings and gives an advisory AHP-style recommendation. The researcher then
creates an explicit `method = "finalize"` model for projection.

Every method stores its figures in the same places, so the same review sequence
works whichever approach you choose. A figure lives beside the table it
summarises rather than in one flat list:

| Location | What it shows |
|---|---|
| `fit_plot` | candidate solutions across the search grid |
| `ModelInfo$plots` | structure of the selected solution |
| `ModelInfo$FitDiagnostics$plots` | how individual training cases sit inside that structure |
| `ProbFit$plots` | membership confidence |
| `Stability$plots` | bootstrap reproducibility |
| `ProjectionFit$plots` | projected cases against the frozen training reference |

The *contents* are method-specific: K-means and Gower/PAM give you a silhouette
profile, Mclust a BIC curve and an uncertainty map, HDBSCAN a persistence
review, LCA item response profiles, and the reduction pipelines a scree plot.
The SOM pipeline keeps its own richer set, including the interactive `aweSOM`
maps in `ModelInfo_SOM$plots`.

# Choosing a method

| Method | Data and geometry | Primary diagnostics | Projection rule | Ideal situation | Main caveat |
|---|---|---|---|---|---|
| SOM + Mclust | Continuous; topology and gradients matter | map fit, posterior, SOM distance, stability | scaling → SOM node → mixture class | phenotype maps and nonlinear gradients | most complex workflow |
| Mclust | Continuous; Gaussian mixtures | BIC/ICL, entropy, posterior, stability | exact mixture posterior | probabilistic groups with different covariance | sensitive to distributional mismatch |
| PCA + Mclust | Correlated continuous measures | scree, mixture fit, posterior, stability | frozen PCA → mixture posterior | clustering a lower-dimensional signal space | PCs can obscure individual variables |
| K-means | Numeric; compact centroid geometry | WSS, silhouette, CH, margin, stability | nearest frozen centroid | simple, interpretable baseline | no probability model |
| PCA + K-means | Correlated numeric measures | scree, silhouette, distance, stability | frozen PCA → nearest centroid | fast reduced-space baseline | inherits PCA and K-means assumptions |
| HDBSCAN | Numeric; irregular or unequal-density groups | persistence, membership, noise, support | nearest eligible training core | clusters plus meaningful noise | conservative projection may label many cases noise |
| Gower + PAM | Mixed numeric/categorical/ordinal | silhouette, medoids, distance, stability | nearest frozen medoid | mixed clinical measures and exemplar participants | coding and variable selection define distance |
| LCA | Categorical response patterns | BIC, entropy, response probabilities, stability | latent-class posterior | symptom or questionnaire response phenotypes | local-independence assumption |
| MCA + Mclust | Nominal categorical measures | inertia, mixture fit, posterior, stability | frozen MCA → mixture posterior | reduced categorical response space | MCA dimensions can be abstract |

# Load the benchmark

```{r}
#| label: LoadPackages
library(SciDataReportR)
library(dplyr)
library(ggplot2)
```

```{r}
#| label: LoadData
data("SimulatedPhenotypeData", package = "SciDataReportR")
data("SimulatedPhenotypeVariableTypes", package = "SciDataReportR")

df_Raw <- SimulatedPhenotypeData
VariableTypes <- SimulatedPhenotypeVariableTypes
df_Revalued <- RevalueData(df_Raw, VariableTypes)$RevaluedData

vars_Continuous <- paste0("Var", 1:12)
vars_Categorical <- paste0("CatVar", 1:3)
vars_Mixed <- c(vars_Continuous, vars_Categorical)
vars_Density <- c("DensityX", "DensityY")

df_Training <- df_Revalued %>%
  dplyr::filter(.data$Cohort == "Training")
df_Projected <- df_Revalued %>%
  dplyr::filter(.data$Cohort == "Projection")
```

::: callout-note
The data file is `SimulatedPhenotypeData`; the codebook is
`SimulatedPhenotypeVariableTypes`. The benchmark contains 320 training and 160
projection participants. `TruthCluster`, `TruthDensityGroup`, and `Cohort` are
never supplied to a clustering function. Truth is revealed only as a teaching
check after fitting.
:::

```{r}
#| label: BenchmarkHelpers
PrintPlots <- function(plots, plot_names = NULL) {
  if (is.null(plot_names)) plot_names <- names(plots)
  for (plot_name in intersect(plot_names, names(plots))) {
    if (inherits(plots[[plot_name]], "ggplot")) print(plots[[plot_name]])
  }
  invisible(NULL)
}

PrintFitMetrics <- function(object, metrics) {
  object$ModelInfo$fit_table %>%
    dplyr::select(dplyr::any_of(c(
      "Model", "Classes", "MinPts", "Epsilon", metrics,
      "ReproducibilityScore", "ahp_index", "Recommended")))
}

CalculateTruthARI <- function(truth, assignment) {
  valid <- !is.na(truth) & !is.na(assignment)
  if (sum(valid) < 2L) return(NA_real_)
  mclust::adjustedRandIndex(truth[valid], assignment[valid])
}

CalculateNoisePerformance <- function(truth, assignment) {
  valid <- !is.na(truth) & !is.na(assignment)
  truth_noise <- as.character(truth[valid]) == "Noise"
  assigned_noise <- assignment[valid] == 0
  dplyr::tibble(
    NoiseSensitivity = mean(assigned_noise[truth_noise]),
    NoiseSpecificity = mean(!assigned_noise[!truth_noise]))
}
```

The 12 numeric measures are three indicators of each of four independent
simulated axes. The default 85% rule therefore retains at least four PCs. Four
balanced truth groups use different combinations of these axes. Separate
`DensityX`/`DensityY` coordinates contain two unequal-density groups and noise;
three categorical variables contain deliberately strong class-dependent
response patterns. That separation is useful for checking implementation; it
is intentionally easier than most real categorical phenotyping problems.

# Continuous workflows

## Mclust

Gaussian mixture models provide posterior membership and allow clusters to
differ in covariance [@scrucca2016].

::: callout-note
**When to use it:** continuous measures with approximately mixture-shaped
groups. **Inspect before naming:** BIC/ICL, entropy, posterior confidence,
small clusters, bootstrap ARI/Jaccard, and projected uncertainty.
:::

```{r}
#| label: MclustReview
df_MclustReview <- CreateMclustClusterModel(
  df_Training, vars_Continuous, k_range = 2:10, model_names = "EEI",
  stability_resamples = 2
)
PrintFitMetrics(
  df_MclustReview,
  c("BIC", "ICL", "Entropy", "MinClusterN", "SizeBalance"))
df_MclustReview$ModelInfo$AHP$recommendation
print(df_MclustReview$fit_plot)
PrintPlots(df_MclustReview$ModelInfo$plots, c("bic", "icl", "entropy"))
PrintPlots(df_MclustReview$Stability$plots)
```

```{r}
#| label: MclustFinalize
df_Mclust <- CreateMclustClusterModel(
  df_Training, vars_Continuous, method = "finalize", final_k = 4,
  final_model = "EEI", ClusterName = "MclustCluster",
  stability_resamples = 2
)
df_MclustProjected <- ProjectMclustCluster(df_Mclust, df_Projected)
PrintPlots(
  df_Mclust$ModelInfo$plots,
  c("separation_map", "classification_uncertainty", "centre_heatmap",
    "profiles"))
PrintPlots(df_Mclust$ModelInfo$FitDiagnostics$plots)
PrintPlots(df_Mclust$ProbFit$plots)
PrintPlots(df_MclustProjected$ProjectionFit$plots)
```

## K-means

K-means is a useful scaled baseline for compact groups [@macqueen1967].

::: callout-note
**When to use it:** continuous features with roughly compact, similarly shaped
groups. **Inspect before naming:** WSS, silhouette, Calinski--Harabasz index,
centroid distance, assignment margin, and bootstrap stability.
:::

```{r}
#| label: KMeansReview
df_KMeansReview <- CreateKMeansClusterModel(
  df_Training, vars_Continuous, k_range = 2:10, nstart = 20,
  stability_resamples = 2
)
PrintFitMetrics(
  df_KMeansReview,
  c("WSS", "Silhouette", "CalinskiHarabasz", "MinClusterN"))
df_KMeansReview$ModelInfo$AHP$recommendation
print(df_KMeansReview$fit_plot)
PrintPlots(
  df_KMeansReview$ModelInfo$plots,
  c("elbow", "silhouette_by_k", "calinski_harabasz"))
PrintPlots(df_KMeansReview$Stability$plots)
```

```{r}
#| label: KMeansFinalize
df_KMeans <- CreateKMeansClusterModel(
  df_Training, vars_Continuous, method = "finalize", final_k = 4,
  nstart = 20, ClusterName = "KMeansCluster", stability_resamples = 2
)
df_KMeansProjected <- ProjectKMeansCluster(df_KMeans, df_Projected)
PrintPlots(
  df_KMeans$ModelInfo$plots,
  c("silhouette", "separation_map", "centre_heatmap", "profiles"))
PrintPlots(df_KMeans$ModelInfo$FitDiagnostics$plots)
PrintPlots(df_KMeans$ProbFit$plots)
PrintPlots(df_KMeansProjected$ProjectionFit$plots)
```

## PCA followed by clustering

PCA concentrates correlated variation into orthogonal dimensions
[@jolliffe2016]. These pipelines refit scaling and PCA inside every stability
bootstrap, then use the frozen training transformation for projection.

::: callout-note
**When to use it:** correlated continuous variables where a reduced signal
space is clearer than the raw feature space. **Inspect before naming:** scree
plot, retained component count/loadings, downstream fit and stability, score
space separation, and projection distance or uncertainty.
:::

```{r}
#| label: PCAMclustReview
df_PCAMclustReview <- CreatePCAMclustClusterModel(
  df_Training, vars_Continuous, k_range = 2:10, model_names = "EEI",
  stability_resamples = 2
)
PrintFitMetrics(
  df_PCAMclustReview,
  c("BIC", "ICL", "Entropy", "MinClusterN"))
df_PCAMclustReview$ModelInfo$AHP$recommendation
ncol(df_PCAMclustReview$PCA$Scores)
print(df_PCAMclustReview$fit_plot)
PrintPlots(df_PCAMclustReview$ModelInfo$plots, c("scree", "bic", "icl"))
PrintPlots(df_PCAMclustReview$Stability$plots)
```

```{r}
#| label: PCAMclustFinalize
df_PCAMclust <- CreatePCAMclustClusterModel(
  df_Training, vars_Continuous, method = "finalize", final_k = 4,
  final_model = "EEI", ClusterName = "PCAMclustCluster",
  stability_resamples = 2
)
df_PCAMclustProjected <- ProjectPCAMclustCluster(df_PCAMclust, df_Projected)
PrintPlots(
  df_PCAMclust$ModelInfo$plots,
  c("scree", "separation_map", "profiles"))
PrintPlots(df_PCAMclust$ModelInfo$FitDiagnostics$plots)
PrintPlots(df_PCAMclust$ProbFit$plots)
PrintPlots(df_PCAMclustProjected$ProjectionFit$plots)
```

```{r}
#| label: PCAKMeansReview
df_PCAKMeansReview <- CreatePCAKMeansClusterModel(
  df_Training, vars_Continuous, k_range = 2:10, nstart = 20,
  stability_resamples = 2
)
PrintFitMetrics(
  df_PCAKMeansReview,
  c("WSS", "Silhouette", "CalinskiHarabasz", "MinClusterN"))
df_PCAKMeansReview$ModelInfo$AHP$recommendation
ncol(df_PCAKMeansReview$PCA$Scores)
print(df_PCAKMeansReview$fit_plot)
PrintPlots(
  df_PCAKMeansReview$ModelInfo$plots,
  c("scree", "elbow", "silhouette_by_k"))
PrintPlots(df_PCAKMeansReview$Stability$plots)
```

```{r}
#| label: PCAKMeansFinalize
df_PCAKMeans <- CreatePCAKMeansClusterModel(
  df_Training, vars_Continuous, method = "finalize", final_k = 4,
  nstart = 20, ClusterName = "PCAKMeansCluster",
  stability_resamples = 2
)
df_PCAKMeansProjected <- ProjectPCAKMeansCluster(df_PCAKMeans, df_Projected)
PrintPlots(
  df_PCAKMeans$ModelInfo$plots,
  c("scree", "silhouette", "separation_map", "profiles"))
PrintPlots(df_PCAKMeans$ModelInfo$FitDiagnostics$plots)
PrintPlots(df_PCAKMeans$ProbFit$plots)
PrintPlots(df_PCAKMeansProjected$ProjectionFit$plots)
```

## HDBSCAN

HDBSCAN is designed for irregular clusters, unequal density, and noise
[@mcinnes2017]. A projected `0` is an out-of-support/noise assignment, not a
new phenotype.

::: callout-note
**When to use it:** density structure is more plausible than centroid geometry
and noise is scientifically meaningful. **Inspect before naming:** persistence,
membership, noise burden, outlier scores, stability, and nearest-core support.
All-noise candidates are valid negative results.
:::

```{r}
#| label: HDBSCANReview
df_HDBSCANReview <- CreateHDBSCANClusterModel(
  df_Training, vars_Density, minPts_range = c(6, 10, 14),
  cluster_selection_epsilon_range = c(0, 0.05), stability_resamples = 2
)
PrintFitMetrics(
  df_HDBSCANReview,
  c("Clusters", "Persistence", "MeanMembershipProbability",
    "NoiseProportion", "MinClusterN"))
df_HDBSCANReview$ModelInfo$AHP$recommendation
print(df_HDBSCANReview$fit_plot)
PrintPlots(df_HDBSCANReview$ModelInfo$plots, c("density_review", "persistence"))
PrintPlots(df_HDBSCANReview$Stability$plots)
```

```{r}
#| label: HDBSCANFinalize
df_HDBSCAN <- CreateHDBSCANClusterModel(
  df_Training, vars_Density, method = "finalize", final_minPts = 10,
  final_cluster_selection_epsilon = 0, ClusterName = "HDBSCANCluster",
  stability_resamples = 2
)
df_HDBSCANProjected <- ProjectHDBSCANCluster(df_HDBSCAN, df_Projected)
PrintPlots(
  df_HDBSCAN$ModelInfo$plots,
  c("separation_map", "cluster_persistence", "profiles"))
PrintPlots(df_HDBSCAN$ModelInfo$FitDiagnostics$plots)
PrintPlots(df_HDBSCAN$ProbFit$plots)
PrintPlots(df_HDBSCANProjected$ProjectionFit$plots)
```

## SOM followed by Mclust

Self-organizing maps preserve local topology before mixture clustering
[@kohonen1982].

::: callout-note
**When to use it:** a map of phenotype gradients is as useful as the final
labels. **Inspect before naming:** map topology, quantization/SOM distance,
node occupancy, posterior confidence, stability, and projected distance flags.
:::

```{r}
#| label: SOMReview
df_SOMReview <- CreateSOMClusterModel(
  df_Training, vars_Continuous, method = "exploratory", k_range = 2:10,
  models = 1, som_xdim = 5, som_ydim = 5, stability_resamples = 2,
  Relabel = FALSE, min_nodes_per_cluster = NULL
)
df_SOMReview$ModelInfo_MClust$fit_table
df_SOMReview$ModelInfo_MClust$AHP$recommendation
df_SOMReview$Stability$summary
print(df_SOMReview$fit_plot)
df_SOMReview$ModelInfo_SOM$plots$Circular
df_SOMReview$ModelInfo_SOM$plots$Line
df_SOMReview$ModelInfo_SOM$plots$Cloud
```

```{r}
#| label: SOMFinalize
df_SOM <- CreateSOMClusterModel(
  df_Training, vars_Continuous, method = "finalize", final_k = 4,
  final_model = 1, som_xdim = 5, som_ydim = 5,
  ClusterName = "SOMCluster", stability_resamples = 2,
  Relabel = FALSE, min_nodes_per_cluster = NULL
)
df_SOMProjected <- ProjectSOMCluster(df_SOM, df_Projected)
PrintPlots(df_SOM$ModelInfo_SOM$plots)
PrintPlots(df_SOM$ModelInfo_SOM$SOMFit$plots)
PrintPlots(df_SOM$ModelInfo_SOM$PhenotypeReference$plots)
PrintPlots(df_SOM$ProbFit$plots)
PrintPlots(df_SOM$Stability$plots)
PrintPlots(
  df_SOMProjected$ProjectionFit$plots,
  c("distance_by_cluster_box", "training_vs_projected_cluster_occupancy",
    "projection_fit_class_bar", "som_grid_map"))
```

# Mixed and categorical workflows

## Gower distance followed by PAM

Gower distance combines mixed feature types, while PAM represents clusters by
observed medoids [@gower1971; @kaufman1990]. Projection uses training ranges,
levels, and medoids rather than recalculating them in the new cohort.

::: callout-note
**When to use it:** numeric, binary, ordinal, and nominal measures must be
clustered together. **Inspect before naming:** candidate silhouette, size
balance, medoids, numeric and categorical profiles, distance/margin, and
bootstrap stability.
:::

```{r}
#| label: GowerPAMReview
df_GowerReview <- CreateGowerPAMClusterModel(
  df_Training, vars_Mixed, k_range = 2:10, stability_resamples = 2
)
PrintFitMetrics(
  df_GowerReview,
  c("Silhouette", "MeanMedoidDistance", "MinClusterN", "SizeBalance"))
df_GowerReview$ModelInfo$AHP$recommendation
print(df_GowerReview$fit_plot)
PrintPlots(
  df_GowerReview$ModelInfo$plots,
  c("silhouette_by_k", "medoid_distance_by_k"))
PrintPlots(df_GowerReview$Stability$plots)
```

```{r}
#| label: GowerPAMFinalize
df_Gower <- CreateGowerPAMClusterModel(
  df_Training, vars_Mixed, method = "finalize", final_k = 4,
  ClusterName = "GowerCluster", stability_resamples = 2
)
df_GowerProjected <- ProjectGowerPAMCluster(df_Gower, df_Projected)
PrintPlots(
  df_Gower$ModelInfo$plots,
  c("silhouette", "gower_map", "profiles",
    "categorical_composition"))
PrintPlots(df_Gower$ModelInfo$FitDiagnostics$plots)
PrintPlots(df_Gower$ProbFit$plots)
PrintPlots(df_GowerProjected$ProjectionFit$plots)
```

## Latent class analysis

LCA models class-specific categorical response probabilities [@linzer2011].

::: callout-note
**When to use it:** symptom, diagnosis, questionnaire, or assay-call items are
the phenotype definition. **Inspect before naming:** BIC, entropy, class size,
response-probability heatmap, posterior confidence, and bootstrap stability.
Low posterior confidence is evidence of weak class separation.
:::

```{r}
#| label: LatentClassReview
df_LCAReview <- CreateLatentClassClusterModel(
  df_Training, vars_Categorical, k_range = 2:10, nrep = 5,
  stability_resamples = 2
)
PrintFitMetrics(
  df_LCAReview,
  c("BIC", "AIC", "Entropy", "MinClassN", "SizeBalance"))
df_LCAReview$ModelInfo$AHP$recommendation
print(df_LCAReview$fit_plot)
PrintPlots(df_LCAReview$ModelInfo$plots, c("bic", "entropy"))
PrintPlots(df_LCAReview$Stability$plots)
```

```{r}
#| label: LatentClassFinalize
df_LCA <- CreateLatentClassClusterModel(
  df_Training, vars_Categorical, method = "finalize", final_k = 4,
  nrep = 5, ClusterName = "LCACluster", stability_resamples = 2
)
df_LCAProjected <- ProjectLatentClassCluster(df_LCA, df_Projected)
PrintPlots(
  df_LCA$ModelInfo$plots,
  c("response_probabilities", "item_profiles", "posterior_map",
    "categorical_composition"))
PrintPlots(df_LCA$ProbFit$plots)
PrintPlots(df_LCAProjected$ProjectionFit$plots)
```

## MCA followed by Mclust

MCA is a dimension-reduction method for nominal categorical measures
[@greenacre2017]. The bootstrap refits MCA before refitting the mixture.

::: callout-note
**When to use it:** several nominal variables form a useful lower-dimensional
category space. **Inspect before naming:** inertia/scree plot, MCA dimensions,
mixture fit, score-space separation, posterior confidence, and stability.
:::

```{r}
#| label: MCAMclustReview
df_MCAMclustReview <- CreateMCAMclustClusterModel(
  df_Training, vars_Categorical, k_range = 2:10, model_names = "EEI",
  stability_resamples = 2
)
PrintFitMetrics(
  df_MCAMclustReview,
  c("BIC", "ICL", "Entropy", "MinClusterN", "SizeBalance"))
df_MCAMclustReview$ModelInfo$AHP$recommendation
print(df_MCAMclustReview$fit_plot)
PrintPlots(df_MCAMclustReview$ModelInfo$plots, c("scree", "bic", "icl"))
PrintPlots(df_MCAMclustReview$Stability$plots)
```

```{r}
#| label: MCAMclustFinalize
df_MCAMclust <- CreateMCAMclustClusterModel(
  df_Training, vars_Categorical, method = "finalize", final_k = 4,
  final_model = "EEI", ClusterName = "MCAMclustCluster",
  stability_resamples = 2
)
df_MCAMclustProjected <- ProjectMCAMclustCluster(df_MCAMclust, df_Projected)
PrintPlots(
  df_MCAMclust$ModelInfo$plots,
  c("scree", "separation_map", "categorical_composition"))
PrintPlots(df_MCAMclust$ModelInfo$FitDiagnostics$plots)
PrintPlots(df_MCAMclust$ProbFit$plots)
PrintPlots(df_MCAMclustProjected$ProjectionFit$plots)
```

# Reveal truth for teaching only

The benchmark truth was not used during fitting. ARI is label-invariant, so
cluster numbers do not need to match the truth labels. Strong simulation
recovery confirms that the examples behave as designed; it is not external or
biological validation.

```{r}
#| label: TruthAgreement
df_TruthAgreement <- dplyr::tibble(
  Method = c(
    "Mclust", "K-means", "PCA + Mclust", "PCA + K-means",
    "Gower + PAM", "LCA", "MCA + Mclust", "SOM + Mclust"),
  TrainingARI = c(
    CalculateTruthARI(df_Training$TruthCluster, df_Mclust$df_with_clusters$MclustCluster),
    CalculateTruthARI(df_Training$TruthCluster, df_KMeans$df_with_clusters$KMeansCluster),
    CalculateTruthARI(df_Training$TruthCluster, df_PCAMclust$df_with_clusters$PCAMclustCluster),
    CalculateTruthARI(df_Training$TruthCluster, df_PCAKMeans$df_with_clusters$PCAKMeansCluster),
    CalculateTruthARI(df_Training$TruthCluster, df_Gower$df_with_clusters$GowerCluster),
    CalculateTruthARI(df_Training$TruthCluster, df_LCA$df_with_clusters$LCACluster),
    CalculateTruthARI(df_Training$TruthCluster, df_MCAMclust$df_with_clusters$MCAMclustCluster),
    CalculateTruthARI(df_Training$TruthCluster, df_SOM$df_with_clusters$SOMCluster)),
  ProjectionARI = c(
    CalculateTruthARI(df_Projected$TruthCluster, df_MclustProjected$ProbFit$individual$Cluster),
    CalculateTruthARI(df_Projected$TruthCluster, df_KMeansProjected$ProbFit$individual$Cluster),
    CalculateTruthARI(df_Projected$TruthCluster, df_PCAMclustProjected$ProbFit$individual$Cluster),
    CalculateTruthARI(df_Projected$TruthCluster, df_PCAKMeansProjected$ProbFit$individual$Cluster),
    CalculateTruthARI(df_Projected$TruthCluster, df_GowerProjected$ProbFit$individual$Cluster),
    CalculateTruthARI(df_Projected$TruthCluster, df_LCAProjected$ProbFit$individual$Cluster),
    CalculateTruthARI(df_Projected$TruthCluster, df_MCAMclustProjected$ProbFit$individual$Cluster),
    CalculateTruthARI(df_Projected$TruthCluster, df_SOMProjected$ProbFit$individual$Cluster))
)
df_TruthAgreement
```

```{r}
#| label: DensityTruthAgreement
df_DensityTruth <- dplyr::tibble(
  Cohort = c("Training", "Projection"),
  ARI = c(
    CalculateTruthARI(
      df_Training$TruthDensityGroup,
      df_HDBSCAN$df_with_clusters$HDBSCANCluster),
    CalculateTruthARI(
      df_Projected$TruthDensityGroup,
      df_HDBSCANProjected$ProbFit$individual$Cluster)),
  NoiseProportion = c(
    mean(df_HDBSCAN$df_with_clusters$HDBSCANCluster == 0, na.rm = TRUE),
    mean(df_HDBSCANProjected$ProbFit$individual$Cluster == 0, na.rm = TRUE))) %>%
  dplyr::bind_cols(dplyr::bind_rows(
    CalculateNoisePerformance(
      df_Training$TruthDensityGroup,
      df_HDBSCAN$df_with_clusters$HDBSCANCluster),
    CalculateNoisePerformance(
      df_Projected$TruthDensityGroup,
      df_HDBSCANProjected$ProbFit$individual$Cluster)))
df_DensityTruth
```

# Compare and interpret profiles

The final label is not the phenotype interpretation. Compare occupancy and
feature distributions across plausible models, then use codebook labels when
describing the profiles.

```{r}
#| label: CompareClusterOccupancy
df_Comparison <- df_Mclust$df_with_clusters %>%
  dplyr::select(.data$.scidr_rowid, .data$MclustCluster) %>%
  dplyr::left_join(
    df_KMeans$df_with_clusters %>%
      dplyr::select(.data$.scidr_rowid, .data$KMeansCluster),
    by = ".scidr_rowid")

df_Comparison %>%
  dplyr::count(.data$MclustCluster, .data$KMeansCluster)
```

```{r}
#| label: PlotClusterProfiles
PlotClusterBoxplot(
  data = df_Mclust$df_with_clusters,
  ClusterVar = "MclustCluster",
  variables = vars_Continuous,
  codebook = VariableTypes,
  Scale = TRUE
)
```

# Reproducibility

# Candidate metrics and SOM + Mclust workflow

All clustering functions have two modes. `method = "exploratory"` compares the
requested 2:10 candidate settings and returns the fit table, candidate audit,
and AHP advisory only. `method = "finalize"` requires the selected setting and
returns assignments, profiles, fit diagnostics, and a reusable projection
model. AHP is advisory; its selected row is retained in the table but is never
highlighted on a figure.

`SizeBalance` is the smallest extracted cluster/class size divided by the
largest (1 is perfectly balanced). `MinClusterN`/`MinClassN` is the smallest
group size. Bootstrap `ARI` measures agreement for the entire partition,
whereas per-cluster Jaccard recovery measures whether each reference phenotype
reappears after refitting. `ReproducibilityScore` is the mean of bootstrap mean
ARI and mean Jaccard recovery; bootstrap success rate is reported separately.

| Method | Candidate metrics |
|---|---|
| Mclust, PCA + Mclust, MCA + Mclust | BIC/ICL/AIC (information criteria), entropy, maximum posterior uncertainty, minimum cluster size, size balance |
| K-means, PCA + K-means | WSS, between-cluster sum of squares, silhouette width, Calinski–Harabasz index, minimum size, size balance |
| Gower + PAM | silhouette width, mean distance to assigned medoid, minimum size, size balance |
| Latent class | BIC, AIC, log likelihood, entropy, minimum class size, size balance |
| HDBSCAN and SOM + HDBSCAN | persistence, noise proportion, membership probability, minimum size, size balance; extracted class count is data-derived |
| SOM + Mclust | LPA AIC/BIC/entropy plus SOM node occupancy and bootstrap stability |

`CreateSOMMclustClusterModel()` is the explicit SOM + Mclust API.
`CreateSOMClusterModel()` remains available for compatibility but is deprecated.
The SOM first freezes Z-score preprocessing and learns map nodes; Mclust then
clusters those node prototypes. A SOM distance is the distance, in this frozen
scaled feature space, between a participant and their best-matching map node.
Lower values mean the participant is better represented by the map; they do
not mean stronger membership in a phenotype. Interpret it against the frozen
training distribution only: distances above the training p95 flag a case that
is poorly represented by the training phenotype space.

# How projection works

Every fitted clustering object freezes its preprocessing and records it in
`Specification`; every projected object reports participant-level diagnostics
in `ProjectionFit$individual`, cohort summaries in `ProjectionFit$summary`,
cluster summaries in `ProjectionFit$by_cluster`, support violations in
`ProjectionFit$out_of_support`, and the applied thresholds in
`ProjectionFit$policy`.

| Workflow | Frozen projection rule | Assignment evidence |
|---|---|---|
| Mclust | Gaussian-mixture `predict()` after training scaling | Posterior probabilities and assigned-component Mahalanobis distance |
| K-means | Nearest frozen centroid | Euclidean centroid distance and second-centroid margin; not a probability |
| PCA/MCA + clustering | Training scaling and frozen reduction, then the fitted downstream model | The downstream model's metric/probability |
| Gower + PAM | Nearest frozen observed medoid using training ranges and levels | Gower medoid distance and runner-up margin |
| Latent class | Frozen class prevalence and item-response probabilities | Posterior class probabilities |
| HDBSCAN | Conservative nearest in-training-cluster support rule | Nearest-core distance and support flag; **not native HDBSCAN prediction** |
| SOM + Mclust | Frozen scaling, SOM BMU, and node mixture probabilities | BMU distance and node-derived posterior probability |
| SOM + HDBSCAN | Frozen scaling, SOM BMU, and node HDBSCAN label | BMU distance and inherited node cluster/noise label; **not native HDBSCAN prediction** |

`ReproducibilityScore` is the mean of bootstrap mean ARI and mean per-cluster
Jaccard recovery. ARI summarizes agreement for the whole partition, whereas
Jaccard asks whether each individual phenotype was recovered. Bootstrap
success rate is reported separately and is not folded into the score.

<details>
<summary>Session information</summary>

```{r}
#| label: Reproducibility
# save.image()
print(sessionInfo())
```

</details>

# References

::: {#refs}
:::
