---
title: "FDR correction scope: matrix-wide vs per-outcome"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FDR correction scope: matrix-wide vs per-outcome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup, message = FALSE}
library(SciDataReportR)
```

The heatmap/matrix family in SciDataReportR (`PlotCorrelationsHeatmap()`,
`PlotPhiHeatmap()`, `PlotChiSqCovar()`, `PlotAnovaRelationshipsMatrix()`,
`PlotMiningMatrix()`, the interaction-effect matrices, and
`PlotDirectionalHeatmaps()`) tests many hypotheses at once, so every one of
these functions applies a false discovery rate (FDR) correction. As of
version 19.15.0 all of that correction runs through one exported helper,
`ApplyFDRCorrection()`, and every function in the family takes an
`fdr_scope` argument that controls *which family of tests* the correction
is computed over:

* `fdr_scope = "matrix"` (the default, and the historical behavior):
  all p-values in the matrix are corrected together as one family.
* `fdr_scope = "per_outcome"`: each outcome's p-values are corrected
  separately. Each function documents which axis is the "outcome" axis in
  its `fdr_scope` parameter documentation.

## `ApplyFDRCorrection()` directly

`ApplyFDRCorrection()` works on a matrix (or data frame) of p-values.
Here is a small made-up example, with predictors as rows and outcomes as
columns:

```{r}
pm <- matrix(
  c(0.005, 0.04, 0.03, 0.01, NA, 0.002),
  nrow = 2,
  dimnames = list(c("pred1", "pred2"), c("out1", "out2", "out3"))
)
pm
```

Matrix-wide correction treats all five finite p-values as one family
(`NA` cells are left untouched):

```{r}
ApplyFDRCorrection(pm, fdr_scope = "matrix")
```

Per-outcome correction adjusts each column on its own. Note how `out3`'s
single test is barely penalized at all, while under matrix-wide correction
it shared the burden of all five tests:

```{r}
ApplyFDRCorrection(pm, fdr_scope = "per_outcome", outcome_margin = 2)
```

## The two scopes in a real analysis

Using `mtcars`, we relate three engine/design predictors to two performance
outcomes with `PlotCorrelationsHeatmap()`. In this function the p-value
matrix has predictors as rows and outcomes as columns, so
`"per_outcome"` corrects within each column (each variable in
`outcome_vars`).

```{r}
res_matrix <- PlotCorrelationsHeatmap(
  mtcars,
  predictor_vars = c("disp", "hp", "drat", "wt"),
  outcome_vars   = c("mpg", "qsec"),
  fdr_scope      = "matrix"
)

res_per_outcome <- PlotCorrelationsHeatmap(
  mtcars,
  predictor_vars = c("disp", "hp", "drat", "wt"),
  outcome_vars   = c("mpg", "qsec"),
  fdr_scope      = "per_outcome"
)
```

The unadjusted p-values are identical by construction:

```{r}
round(res_matrix$p$p, 5)          # res$p is an alias for res$Unadjusted
all.equal(res_matrix$p$p, res_per_outcome$p$p)
```

The FDR-adjusted p-values differ (`res$p_fdr` is an alias for
`res$FDRCorrected`):

```{r}
round(res_matrix$p_fdr$p, 5)       # one family of 8 tests
round(res_per_outcome$p_fdr$p, 5)  # two families of 4 tests each
```

Why do they differ? The Benjamini-Hochberg adjustment scales each p-value
by (number of tests in the family) / (rank within the family). Under
`"matrix"` scope, `mpg` and `qsec` tests compete in one ranking of 8
tests. Under `"per_outcome"` scope, the 4 `mpg` tests are ranked only
against each other, and likewise for `qsec` - so a strong predictor of a
"mostly null" outcome is no longer penalized for significant tests that
belong to a different outcome, and vice versa.

The FDR-starred heatmaps reflect the same difference:

```{r, fig.show = "hold"}
res_matrix$p_fdr$plot
res_per_outcome$p_fdr$plot
```

## Choosing a scope

* Use `"matrix"` (the default) when you view the whole screen as one
  discovery exercise - "which of all these relationships are real?" This
  is the more conservative choice for the individual outcome and is what
  every SciDataReportR version before 19.15.0 did.
* Use `"per_outcome"` when each outcome is a separate scientific question
  and you want the error rate controlled within each outcome's own set of
  tests - a common convention when outcomes will be reported in separate
  results sections or figures.

Whichever you choose, report it: the two scopes answer different
questions, and the choice should be made before looking at the results.
