---
title: "Getting started with SciDataReportR"
vignette: >
  %\VignetteIndexEntry{Getting started with SciDataReportR}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 2
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

SciDataReportR helps researchers turn labelled scientific and clinical data into reproducible exploratory analyses, statistical comparisons, scientific visualizations, and report-ready outputs.

This article is adapted from the SciDataReportR R/Medicine workflow narrative. It follows the common path from a raw researcher-native data frame to metadata-aware reporting and reusable analysis outputs.

The examples use the current workflow-oriented function names. Older names such as `CreatePCATable()`, `CreateZScorePlot()`, and `Make_DataDictionary()` remain available as compatibility aliases.

## Why this workflow matters

Scientific and clinical datasets often contain many variables, labels, recoding rules, caveats, missing values, and potential relationships. Before formal statistical modeling, researchers need to understand the structure of the data, identify quality issues, inspect distributions, screen associations, and decide whether covariates or dimensionality reduction are needed.

SciDataReportR organizes that work into repeatable steps:

1. Import data and create metadata.
2. Recode and relabel variables.
3. Profile missingness and distributions.
4. Create summary and comparison tables.
5. Screen associations across variable families.
6. Build dimensionality reduction or projection workflows.
7. Save reproducible, report-ready outputs.

## Load data

Most SciDataReportR workflows start with a data frame and, when available, a variable-type table or codebook.

```{r}
#| label: LoadPackages
library(SciDataReportR)
```

```{r}
#| label: LoadData
data("SampleData")
data("SampleVariableTypes")

df_Raw <- SampleData
VariableTypes <- SampleVariableTypes
```

::: callout-note
The examples use the raw `SampleData` dataset and its shipped `SampleVariableTypes` codebook. The raw `age` variable contains `999` as a sentinel for missing data; `RevalueData()` uses the codebook to convert those values to `NA` before any summaries, plots, or models are created.
:::

For a project dataset, the starting point might look like this:

```{r}
#| label: ProjectDataExample
#| eval: false
df_Clinical <- readr::read_csv(here::here("data", "clinical_data.csv"))
VariableTypes_Clinical <- readr::read_csv(
  here::here("data", "variable_types.csv")
)
```

## Create variable metadata

Variable type templates make the analysis plan explicit. They help distinguish continuous, categorical, binary, ordinal, outcome, covariate, and feature variables before plotting or testing.

```{r}
#| label: CreateVariableMetadata
VariableTypes_Template <- CreateVariableTypesTemplate(df_Raw)
```

`CreateVariableTypesTemplate()` is useful when a dataset does not yet have a codebook. For the remaining examples, use the shipped `SampleVariableTypes` object because it already contains the labels, recodes, and missing-value rules for `SampleData`.

Data dictionaries summarize the available variables and make the dataset easier to review with collaborators.

```{r}
#| label: ReviewDataDictionary
df_DataDictionary <- MakeDataDictionary(df_Raw)
FormattedDataDictionary(df_Raw)
```

## Recode and relabel data

When a variable-type table contains recoding or label information, use it to produce an analysis-ready data frame while preserving scientific meaning.

```{r}
#| label: RevalueSampleData
df_Revalued <- RevalueData(df_Raw, VariableTypes)$RevaluedData
```

## Inspect missingness

Missing data patterns are one of the first bottlenecks in scientific analysis. They affect which variables are interpretable, which models are feasible, and whether downstream comparisons are biased.

```{r}
#| label: PlotMissingness
PlotMissingData(df_Revalued, Relabel = TRUE)
```

## Create report-ready summaries

Table 1 style summaries help reviewers and collaborators understand the cohort before detailed analyses.

```{r}
#| label: CreateTableOne
MakeTable1(df_Revalued, TreatOrdinalAs = "Continuous")
```

Continuous and categorical profiling functions help researchers inspect variable distributions without hand-building repeated plots.

```{r}
#| label: ProfileDistributions
vars_Continuous <- getNumVars(df_Revalued, Ordinal = FALSE)
vars_Categorical <- getCatVars(df_Revalued)

CreateSummaryTable(df_Revalued, vars_Continuous, Relabel = TRUE)
PlotContinuousDistributions(df_Revalued, vars_Continuous[1:12], ncol = 3)
PlotCategoricalDistributions(df_Revalued, vars_Categorical)
```

## Screen associations

SciDataReportR supports both focused plots and matrix-style screening. This helps researchers identify relationships, candidate covariates, and patterns that may need dimensionality reduction.

```{r}
#| label: PlotFocusedAssociations
PlotAssociations(df_Revalued, "age", "Adiponectin")
PlotAssociations(df_Revalued, "Diagnosis", "Ab_42")
```

Correlation heatmaps return structured objects that can be reused downstream.

```{r}
#| label: PlotCorrelationHeatmap
correlation_result <- PlotCorrelationsHeatmap(
  df_Revalued,
  xVars = vars_Continuous[1:5],
  yVars = vars_Continuous[20:40],
  method = "pearson",
  covars = NULL,
  Relabel = TRUE,
  Ordinal = FALSE
)

correlation_result$Unadjusted$plot
```

The output of `PlotCorrelationsHeatmap()` can be passed to `add_r_and_stars()`, and the returned plot can use `geom_starcaption()` to explain significance stars.

```{r}
#| label: AnnotateCorrelationHeatmap
add_r_and_stars(correlation_result) + geom_starcaption()
```

## Compare groups

Group comparisons are a central part of clinical and life science reporting. `MakeComparisonTable()` provides report-ready summaries with optional covariates, effect sizes, and pairwise contrasts.

```{r}
#| label: CompareDiagnosisGroups
MakeComparisonTable(
  DataFrame = df_Revalued,
  CompVariable = "Diagnosis",
  Variables = c("age", "tau", "p_tau"),
  AddEffectSize = TRUE
)
```

For high-dimensional group signals, z-score plots provide a compact visual summary across many variables.

```{r}
#| label: PlotGroupZScores
vars_LabMeasures <- vars_Continuous[10:60]

PlotZScore(
  df_Revalued,
  TargetVar = "Diagnosis",
  Variables = vars_LabMeasures,
  sort = FALSE
)
```

## Build reusable dimensionality reduction workflows

PCA workflows can produce scree plots, loading summaries, and reusable projection objects. This is useful when a cohort-level transformation should be applied to future datasets.

```{r}
#| label: CreatePCAObject
pca_object <- CreatePCAObject(
  df_Revalued,
  vars_LabMeasures,
  minThresh = 0.75,
  scale = TRUE,
  center = TRUE
)

pca_object$p_scree
pca_object$Lollipop
```

Projection functions make the fit/apply split explicit:

```{r}
#| label: ProjectPCAExample
#| eval: false
projected_scores <- ProjectPCA(new_data, pca_object)
```

## Reproducibility

SciDataReportR workflows are designed to make repeated analysis steps visible and reusable. For reports, include session information and preserve the variable-type table, codebook, and any fit objects used for projections.

```{r}
#| label: Reproducibility
# save.image(here::here("results", "getting-started.RData"))
print(sessionInfo())
```

## Next steps

Use the reference index by workflow family:

- Data setup, metadata, and codebooks.
- Preprocessing and data quality.
- Statistical comparison workflows.
- Association, regression, and interaction workflows.
- Visualization functions.
- Dimensionality reduction, projection, and clustering.
- Longitudinal and temporal workflows.

Future article additions should expand this starter flow into focused workflows for codebook harmonization, visualization galleries, PCA/projection, SOM projection, normative T-scores, and longitudinal transitions.
