---
title: "Creating comparison tables with MakeComparisonTable()"
description: "Generate publication-ready comparison tables with effect sizes, covariate adjustment, and pairwise comparisons."
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Creating comparison tables with MakeComparisonTable()}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

`MakeComparisonTable()` creates publication-ready summary tables for comparing groups across continuous and categorical variables.

This vignette demonstrates:

* Basic group comparisons
* Adding effect sizes
* Parametric versus nonparametric testing
* Covariate adjustment
* Pairwise comparisons
* Reference-group comparisons
* Automatic use of variable labels

## Load packages

```{r}
library(SciDataReportR)
library(dplyr)
```

## Load example data

SciDataReportR includes an example dataset and codebook.

```{r}
data("SampleData")
data("SampleVariableTypes")

RevaluedObj <- RevalueData(
  SampleData,
  SampleVariableTypes
) 
df_Revalued <- RevaluedObj$RevaluedData
```


## Basic comparison table

The most common use case is comparing groups across a set of variables.

Here we compare diagnostic groups.

```{r}
tbl_basic <- MakeComparisonTable(
  df_Revalued,
  CompVariable = "Diagnosis",
  Variables = c(
    "age",
    "sex",
    "Genotype",
    "AXL",
    "Calbindin",
    "Ferritin",
    "MMP7"
  )
)

tbl_basic
```

The resulting table contains:

* Descriptive statistics
* Group comparison tests
* P-values
* Automatically applied variable labels

## Adding effect sizes

Effect sizes provide information about the magnitude of group differences.

```{r}
tbl_effectsize <- MakeComparisonTable(
  df_Revalued,
  CompVariable = "Diagnosis",
  Variables = c(
    "age",
    "sex",
    "Genotype",
    "AXL",
    "Calbindin",
    "Ferritin",
    "MMP7"
  ),
  AddEffectSize = TRUE
)

tbl_effectsize
```

When reporting group differences, effect sizes should generally be interpreted alongside p-values.



## Nonparametric analyses

Nonparametric testing uses rank-based methods that are more robust to non-normal distributions.

```{r}
tbl_nonparametric <- MakeComparisonTable(
  df_Revalued,
  CompVariable = "Diagnosis",
  Variables = c(
    "age",
    "sex",
    "Genotype",
    "AXL",
    "Calbindin",
    "Ferritin",
    "MMP7"
  ),
  Parametric = FALSE,
  AddEffectSize = TRUE
)

tbl_nonparametric
```


## Including Covariates

Age is frequently included as a covariate in biomedical analyses.

The example below evaluates group differences after accounting for age.

```{r}
tbl_covariate <- MakeComparisonTable(
  df_Revalued,
  CompVariable = "Diagnosis",
  Variables = c(
    "sex",
    "Genotype",
    "AXL",
    "Calbindin",
    "Ferritin",
    "MMP7"
  ),
  Covariates = "age",
  AddEffectSize = TRUE
)

tbl_covariate
```

Covariate adjustment can help determine whether observed group differences persist after controlling for potential confounding factors.

## Pairwise comparisons

When comparing more than two groups, pairwise comparisons can identify which groups differ from one another.


```{r}
#| column: screen
tbl_pairwise <- MakeComparisonTable(
  df_Revalued,
  CompVariable = "Genotype",
  Variables = c(
    "Diagnosis",
    "age",
    "sex",
    "AXL",
    "Calbindin",
    "Ferritin",
    "MMP7"
  ),
  AddPairwise = TRUE,
  AddEffectSize = TRUE
)

tbl_pairwise
```

You can also choose a referent and compare only to that

```{r}
tbl_pairwise_referent <- MakeComparisonTable(
  df_Revalued,
  CompVariable = "Genotype",
  Variables = c(
    "Diagnosis",
    "age",
    "sex",
    "AXL",
    "Calbindin",
    "Ferritin",
    "MMP7"
  ),
  AddPairwise = TRUE,
  AddEffectSize = TRUE,
  Referent = "E3E3"
)

tbl_pairwise_referent
```


## Summary

`MakeComparisonTable()` supports:

* Descriptive statistics
* Parametric analyses
* Nonparametric analyses
* Effect sizes
* Covariate adjustment
* Pairwise testing
* Reference-group comparisons
* Automatic variable labeling

## Related functions

* `RevalueData()`
* `CreateCodebook()`
* `PlotBoxPlot()`
* `PlotVolcanoEffects()`
* `CompareDatasets()`

## Session information

```{r}
sessionInfo()
```
