---
title: "Screening Outcomes and Predictors with UnivariateRegressionTable()"
description: "Create label-aware univariate regression tables across multiple outcomes and predictors."
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Screening Outcomes and Predictors with UnivariateRegressionTable()}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

`UnivariateRegressionTable()` screens outcome-predictor relationships and returns a report-facing table, a detail table, fitted models, and metadata.

Use it when you want to quickly ask:

* Which predictors are associated with one or more outcomes?
* Do associations persist after covariate adjustment?
* Are the displayed labels and logistic event levels what I think they are?

The function now chooses the model family automatically:

* Numeric outcomes use linear regression.
* Two-level factor, character, or logical outcomes use logistic regression.
* Logistic estimates are exponentiated by default and reported as odds ratios.

## Load packages

```{r}
#| label: LoadPackages
library(SciDataReportR)
library(dplyr)
```

## Load example data

The examples use `SampleData` and `SampleVariableTypes`, then apply labels and recoding with `RevalueData()`.

```{r}
#| label: LoadData
data("SampleData")
data("SampleVariableTypes")

RevaluedObj <- RevalueData(
  SampleData,
  SampleVariableTypes
)

df_Revalued <- RevaluedObj$RevaluedData
```

## Basic linear screening

This example screens several labelled biomarkers against two predictors.

```{r}
#| label: DefineLinearVariables
vars_Outcomes <- c(
  "Calbindin",
  "Ferritin",
  "MMP7",
  "Sortilin"
)

vars_Predictors <- c(
  "Diagnosis",
  "age"
)
```

```{r}
#| label: FitLinearScreen
Reg_Obj_Un <- UnivariateRegressionTable(
  Data = df_Revalued,
  OutcomeVars = vars_Outcomes,
  PredictorVars = vars_Predictors,
  Standardize = TRUE
)
```

## Formatted and detailed tables

`FormattedTable` is the report-facing table.

```{r}
#| label: DisplayFormattedTable
Reg_Obj_Un$FormattedTable
```

`LargeTable` keeps more of the model detail and is useful for QC.

```{r}
#| label: DisplayLargeTable
Reg_Obj_Un$LargeTable
```

## Covariate adjustment

Use `Covars` when each screen should adjust for the same covariates.

```{r}
#| label: FitCovariateScreen
Reg_Obj_Un_Covar <- UnivariateRegressionTable(
  Data = df_Revalued,
  OutcomeVars = vars_Outcomes,
  PredictorVars = vars_Predictors,
  Covars = "sex",
  Standardize = TRUE
)

Reg_Obj_Un_Covar$FormattedTable
```

## Many predictors for one outcome

The same function also works when the question is one outcome against many candidate predictors.

```{r}
#| label: FitManyPredictors
Reg_Obj_ManyPredictors <- UnivariateRegressionTable(
  Data = df_Revalued,
  OutcomeVars = "Sortilin",
  PredictorVars = c("Diagnosis", "age", "Calbindin", "Ferritin", "MMP7"),
  Standardize = TRUE
)

Reg_Obj_ManyPredictors$FormattedTable
```

## Logistic outcomes

Binary outcomes use logistic regression automatically. By default, estimates are exponentiated and shown as odds ratios.

```{r}
#| label: FitLogisticScreen
Reg_Obj_Logistic <- UnivariateRegressionTable(
  Data = df_Revalued,
  OutcomeVars = "Diagnosis",
  PredictorVars = c("age", "Calbindin", "Ferritin", "MMP7"),
  Standardize = TRUE
)

Reg_Obj_Logistic$FormattedTable
```

Always check the metadata for logistic models. It records the reference level and event level.

```{r}
#| label: InspectLogisticMetadata
Reg_Obj_Logistic$Metadata$Outcomes
```

## Access fitted models

Fitted models are returned by outcome, then predictor.

```{r}
#| label: InspectModelObjects
names(Reg_Obj_Un$ModelSummaries)
names(Reg_Obj_ManyPredictors$ModelSummaries$Sortilin)

summary(Reg_Obj_ManyPredictors$ModelSummaries$Sortilin$Ferritin)
```

## Reproducibility

```{r}
#| label: Reproducibility
# save.image("UnivariateRegressionTable_workspace.RData")
print(sessionInfo())
```
