---
title: "Forest Plots from Univariate Regression Tables"
description: "Visualize estimates from UnivariateRegressionTable() with plotForestFromTable()."
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Forest Plots from Univariate Regression Tables}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

`plotForestFromTable()` visualizes the estimates and confidence intervals from a `UnivariateRegressionTable()` object.

The default layout is:

* facets = outcomes
* rows = predictors or predictor levels
* x-axis = regression estimate

Use `Flip = TRUE` when it is easier to read outcomes as rows and predictors as facets.

## Load packages

```{r}
#| label: LoadPackages
library(SciDataReportR)
library(dplyr)
```

## Load example data

The examples use `SampleData` and `SampleVariableTypes`, then apply labels and recoding with `RevalueData()`.

```{r}
#| label: LoadData
data("SampleData")
data("SampleVariableTypes")

RevaluedObj <- RevalueData(
  SampleData,
  SampleVariableTypes
)

df_Revalued <- RevaluedObj$RevaluedData
```

## Many outcomes and predictors

This is a common screening pattern: several outcomes tested against several predictors.

```{r}
#| label: FitOutcomeScreen
vars_Outcomes <- c(
  "Calbindin",
  "Ferritin",
  "MMP7",
  "Sortilin"
)

vars_Predictors <- c(
  "Diagnosis",
  "age"
)

Reg_Obj_Outcomes <- UnivariateRegressionTable(
  Data = df_Revalued,
  OutcomeVars = vars_Outcomes,
  PredictorVars = vars_Predictors,
  Standardize = TRUE
)
```

## Default forest plot

The default plot keeps outcomes as facets and terms as rows.

```{r}
#| label: PlotDefaultForest
plotForestFromTable(Reg_Obj_Outcomes)
```

Black points indicate `p < 0.05`; grey points are not significant. The dashed vertical line is the null value for linear estimates.

## Flip outcomes and predictors

When there are many outcomes and a small predictor set, `Flip = TRUE` is often easier to read.

```{r}
#| label: PlotFlippedForest
plotForestFromTable(
  Reg_Obj_Outcomes,
  Flip = TRUE
)
```

With `Flip = TRUE`, predictors or predictor levels become facets, and outcomes become rows.

## Many predictors, one outcome

For one outcome and many predictors, the default layout is usually already compact.

```{r}
#| label: FitPredictorScreen
Reg_Obj_Predictors <- UnivariateRegressionTable(
  Data = df_Revalued,
  OutcomeVars = "Sortilin",
  PredictorVars = c("Diagnosis", "age", "Calbindin", "Ferritin", "MMP7"),
  Standardize = TRUE
)
```

```{r}
#| label: PlotManyPredictors
plotForestFromTable(Reg_Obj_Predictors)
```

## Labels

Forest plot facets and rows use labels inherited from `SampleVariableTypes` through `RevalueData()`. If labels are missing, the plot falls back to the variable names.

```{r}
#| label: InspectPlotLabels
Reg_Obj_Outcomes$Metadata$Outcomes
```

Categorical and dichotomous predictors are shown with the modeled level, for example `Diagnosis : Impaired`.

## Logistic models

`UnivariateRegressionTable()` uses logistic regression automatically for two-level outcomes and reports odds ratios by default.

```{r}
#| label: FitLogisticScreen
Reg_Obj_Logistic <- UnivariateRegressionTable(
  Data = df_Revalued,
  OutcomeVars = "Diagnosis",
  PredictorVars = c("age", "Calbindin", "Ferritin", "MMP7"),
  Standardize = TRUE
)

Reg_Obj_Logistic$FormattedTable
```

```{r}
#| label: InspectLogisticMetadata
Reg_Obj_Logistic$Metadata$Outcomes
```

```{r}
#| label: PlotLogisticForest
plotForestFromTable(Reg_Obj_Logistic)
```

For logistic models, check the event level in `Metadata$Outcomes` before interpreting odds ratios. Mixed linear and logistic forest plots can be visually convenient, but they combine different effect scales.

## Reproducibility

```{r}
#| label: Reproducibility
# save.image("plotForestFromTable_workspace.RData")
print(sessionInfo())
```
