---
title: "Exploratory discovery with PlotMiningMatrix()"
description: "Rapidly identify patterns, anomalies, covariates, and candidate hypotheses across an entire dataset."
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Exploratory discovery with PlotMiningMatrix()}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

`PlotMiningMatrix()` is designed for exploratory data mining and hypothesis generation.

Rather than requiring separate correlation analyses, ANOVA models, and chi-square tests, the function automatically identifies appropriate statistical relationships based on variable type and summarizes the results in a single visualization.

This function is particularly useful for:

* Identifying unexpected relationships
* Detecting potential confounders and covariates
* Discovering clusters of related variables
* Generating hypotheses for future analyses
* Performing exploratory quality control
* Prioritizing variables for downstream modeling

Unlike a traditional correlation matrix, `PlotMiningMatrix()` can simultaneously evaluate:

* Continuous vs continuous relationships
* Categorical vs continuous relationships
* Categorical vs categorical relationships

using appropriate statistical methods for each comparison.

## Load packages

```{r}
library(SciDataReportR)
library(dplyr)
library(plotly)
```

## Load example data

```{r}
data("SampleData")
data("SampleVariableTypes")

RevaluedObj <- RevalueData(
  SampleData,
  SampleVariableTypes
)

df_Revalued <- RevaluedObj$RevaluedData
```

## Create a mining matrix

For exploratory mining it is often useful to evaluate nearly the entire dataset simultaneously.

```{r}
vars_mining <- names(df_Revalued)

length(vars_mining)
```

```{r}
#| column: screen
#| fig-height: 12
#| fig-width: 18
MiningObj <- PlotMiningMatrix(
  Data = df_Revalued,
  OutcomeVars = vars_mining
)

MiningObj$Unadjusted$plot
```

The resulting visualization summarizes hundreds or thousands of statistical relationships simultaneously.

Color intensity represents effect size magnitude, while symbol size and shape represent statistical significance.

## Understanding the statistical tests

`PlotMiningMatrix()` automatically selects appropriate analyses based on variable type.

| Variable Types | Statistical Test |
|---------------|------------------|
| Continuous vs Continuous | Correlation |
| Categorical vs Continuous | ANOVA |
| Categorical vs Categorical | Chi-square |

This allows mixed datasets to be explored without manually specifying separate analyses.

## Interpreting the mining matrix

The mining matrix is best viewed as a discovery tool rather than a confirmatory analysis.

Questions it can help answer include:

* Which biomarkers appear related to diagnosis?
* Which demographic variables may act as covariates?
* Which variables tend to cluster together?
* Are there unexpected associations worth investigating?
* Are there patterns suggesting data quality issues?

Large effect sizes combined with strong significance often represent promising targets for follow-up analyses.

## Parametric analyses

By default, parametric analyses are used.

```{r}
#| column: screen
#| fig-height: 12
#| fig-width: 18
#| 
MiningParametric <- PlotMiningMatrix(
  Data = df_Revalued,
  OutcomeVars = vars_mining,
  Parametric = TRUE
)

MiningParametric$Unadjusted$plot
```

For continuous variables this uses Pearson correlations and parametric group comparisons.

## Nonparametric analyses

For skewed variables or datasets containing substantial outliers, nonparametric methods may be preferable.

```{r}
#| column: screen
#| fig-height: 12
#| fig-width: 18


MiningNonParametric <- PlotMiningMatrix(
  Data = df_Revalued,
  OutcomeVars = vars_mining,
  Parametric = FALSE
)

MiningNonParametric$Unadjusted$plot
```

This switches continuous-variable analyses to Spearman correlations and uses nonparametric group comparisons where applicable.

## Unadjusted significance

The default plot displays significance based on unadjusted p-values.

```{r}
#| column: screen
#| fig-height: 12
#| fig-width: 18
MiningObj$Unadjusted$plot
```

This view is useful for exploratory analyses and hypothesis generation.

## FDR-corrected significance

Because large mining matrices may involve hundreds or thousands of statistical tests, multiple-testing correction is often recommended.

```{r}
#| column: screen
#| fig-height: 12
#| fig-width: 18
MiningObj$FDRCorrected$plot
```

The FDR-corrected visualization highlights relationships that remain significant after accounting for multiple comparisons.

In exploratory studies, investigators often review both unadjusted and FDR-corrected results.

## Examining the underlying results

The statistical results are stored in the returned object.

```{r}
head(
  MiningObj$Unadjusted$PvalTable
)
```

The table includes:

* Variables involved in the association
* Effect size estimates
* Raw p-values
* FDR-adjusted p-values
* Statistical test type

## Interactive exploration

Large mining matrices can be easier to interpret interactively.

```{r}
#| column: screen
#| fig-height: 12
#| fig-width: 18
#| 
ggplotly(
  MiningObj$Unadjusted$plot
)
```

Interactive plots allow users to:

* Hover over relationships
* Zoom into regions of interest
* Explore large datasets more efficiently

Interactive exploration is particularly useful for large omics and clinical datasets.

## Identifying potential covariates

One common use of `PlotMiningMatrix()` is identifying variables that may need to be included as covariates in downstream analyses.

For example:

* Age may be associated with multiple biomarkers.
* Education may be associated with cognitive outcomes.
* Site may be associated with assay measurements.

Variables showing broad associations throughout the mining matrix often deserve consideration as potential confounders.


## Targeted mining

Although `PlotMiningMatrix()` can evaluate nearly an entire dataset, it can also be restricted to a subset of variables.

```{r}
selected_vars <- c(
  "Diagnosis",
  "age",
  "sex",
  "AXL",
  "Calbindin",
  "Ferritin",
  "MMP7"
)

MiningSubset <- PlotMiningMatrix(
  Data = df_Revalued,
  OutcomeVars = selected_vars
)

MiningSubset$Unadjusted$plot
```

This approach is useful when focusing on a particular biological pathway or clinical domain.

## Recommended workflow

A common SciDataReportR workflow is:

1. Use `PlotMiningMatrix()` to identify potentially interesting relationships.
2. Review FDR-corrected results.
3. Identify potential covariates.
4. Visualize significant findings using `plotSigAssociations()`.
5. Perform targeted follow-up analyses using:
   * `MakeComparisonTable()`
   * `PlotCorrelationsHeatmap()`
   * Regression models
   * Longitudinal analyses

## Summary

`PlotMiningMatrix()` provides a rapid overview of relationships within a dataset and is particularly useful for exploratory analyses and hypothesis generation.

Key features include:

* Automatic test selection
* Mixed continuous and categorical variables
* Effect size visualization
* FDR correction
* Identification of candidate covariates
* Interactive exploration
* Follow-up visualization workflows

## Related functions

* `PlotCorrelationsHeatmap()`
* `PlotAnovaRelationshipsMatrix()`
* `PlotChiSqCovar()`
* `plotSigAssociations()`
* `MakeComparisonTable()`

## Session information

```{r}
sessionInfo()
```
