---
title: "Exploring associations with PlotCorrelationsHeatmap()"
description: "Create correlation heatmaps, perform FDR correction, adjust for covariates, and investigate significant associations."
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Exploring associations with PlotCorrelationsHeatmap()}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

`PlotCorrelationsHeatmap()` computes correlations or partial correlations and visualizes the results as a heatmap.

This vignette demonstrates:

* Correlations among biomarkers
* Correlations between biomarkers and clinical variables
* Pearson versus Spearman correlations
* False discovery rate (FDR) correction
* Covariate-adjusted correlations
* Annotating heatmaps with correlation coefficients and significance stars
* Interactive exploration using plotly
* Follow-up visualization of significant correlations

## Load packages

```{r}
library(SciDataReportR)
library(dplyr)
library(plotly)
```

## Load example data

SciDataReportR includes an example dataset and variable type specification file.

```{r}
data("SampleData")
data("SampleVariableTypes")

RevaluedObj <- RevalueData(
  SampleData,
  SampleVariableTypes
)

df_Revalued <- RevaluedObj$RevaluedData
```

## Select biomarkers

For this vignette we will examine relationships among a panel of biomarkers.

```{r}
biomarkers <- c(
  "AXL",
  "Calbindin",
  "Ferritin",
  "MMP7",
  "MMP10",
  "NFL",
  "GFAP",
  "IL6",
  "VCAM1",
  "ICAM1",
  "CRP",
  "TNFa",
  "MCP1",
  "YKL40",
  "Tau"
)
```

## Correlations among biomarkers

When only `xVars` are supplied, a square correlation matrix is generated.

```{r}
CorrObj <- PlotCorrelationsHeatmap(
  Data = df_Revalued,
  xVars = biomarkers
)

CorrObj$Unadjusted$plot
```

Each tile represents the correlation coefficient between two biomarkers. Positive and negative correlations are represented by opposite ends of the color scale.

## Correlations between biomarkers and clinical variables

A rectangular correlation matrix can be created by supplying both `xVars` and `yVars`.

```{r}
clinical_vars <- c(
  "age",
  "Insulin"
)
```

```{r}
CorrObj_Clinical <- PlotCorrelationsHeatmap(
  Data = df_Revalued,
  xVars = biomarkers,
  yVars = clinical_vars
)

CorrObj_Clinical$Unadjusted$plot
```

This approach is useful for screening biomarker-clinical relationships.

## Pearson versus Spearman correlations

The choice of correlation method depends on the characteristics of the data.

| Method | Recommended Use |
|----------|----------|
| Pearson | Approximately normal variables with linear relationships |
| Spearman | Skewed variables, outliers, or monotonic relationships |
| Kendall | Small samples or ordinal variables |

### Pearson correlations

Pearson correlation measures linear relationships between variables and is most appropriate when variables are approximately normally distributed.

```{r}
PearsonObj <- PlotCorrelationsHeatmap(
  Data = df_Revalued,
  xVars = biomarkers,
  method = "pearson"
)

PearsonObj$Unadjusted$plot
```

### Spearman correlations

Spearman correlation is rank-based and is often preferred for biomarker data because it is more robust to skewed distributions and outliers.

```{r}
SpearmanObj <- PlotCorrelationsHeatmap(
  Data = df_Revalued,
  xVars = biomarkers,
  method = "spearman"
)

SpearmanObj$Unadjusted$plot
```

Researchers frequently compare both methods to determine whether findings are sensitive to distributional assumptions.

## Unadjusted versus FDR-corrected significance

`PlotCorrelationsHeatmap()` automatically computes both unadjusted and FDR-corrected p-values.

### Unadjusted significance

```{r}
CorrObj$Unadjusted$plot
```

### FDR-corrected significance

```{r}
CorrObj$FDRCorrected$plot
```

When evaluating many correlations simultaneously, false discovery rate correction is recommended to reduce the likelihood of false positive findings.

## Adding correlation coefficients and significance stars

The default heatmaps display significance stars. Additional annotations can be added using `add_r_and_stars()`.

### Raw p-values

```{r}
add_r_and_stars(
  CorrObj,
  star_from = "raw"
)
```

### FDR-corrected p-values

```{r}
add_r_and_stars(
  CorrObj,
  star_from = "fdr"
)
```

These annotated heatmaps are useful for presentations and publications.

## Adjusting for covariates

Partial correlations can be estimated by specifying one or more covariates.

The example below adjusts all correlations for age.

```{r}
CorrObj_AgeAdjusted <- PlotCorrelationsHeatmap(
  Data = df_Revalued,
  xVars = biomarkers,
  covars = "age"
)

CorrObj_AgeAdjusted$FDRCorrected$plot
```

When covariates are supplied, variables are residualized prior to correlation testing. This can help account for potential confounding factors.

## Interactive heatmaps

Heatmaps can be converted into interactive visualizations using plotly.

```{r}
ggplotly(
  CorrObj$FDRCorrected$plot
)
```

Interactive plots allow users to:

* Hover over cells to inspect values
* Zoom into regions of interest
* Explore large correlation matrices

## Investigating significant correlations

Significant correlations can be examined individually using `plotSigCorrelations()`.

```{r}
SigPlots <- plotSigCorrelations(
  DataFrame = df_Revalued,
  CorrelationHeatmapObject = CorrObj
)
```

The resulting object contains a list of scatterplots for significant associations.

```{r}
length(SigPlots)
```

These plots can be combined into a single figure.

```{r}
#| column: screen
#| fig-height: 12
#| fig-width: 12
AssemblePlots(
  SigPlots,
  LegendPosition = "none",
  ncol = 3
)
```

This workflow provides a convenient way to visually inspect the relationships underlying significant heatmap cells.

## Accessing correlation results

The returned object stores both correlation estimates and p-values.

### Unadjusted results

```{r}
head(
  CorrObj$Unadjusted$r
)
```

### FDR-corrected results

```{r}
head(
  CorrObj$FDRCorrected$p
)
```

These results can be exported for reporting or used in downstream analyses.

## Summary

`PlotCorrelationsHeatmap()` provides a comprehensive workflow for exploratory association analysis, including:

* Pearson, Spearman, and Kendall correlations
* Partial correlations using covariate adjustment
* Automatic FDR correction
* Label-aware visualization
* Interactive heatmaps
* Correlation coefficient annotation
* Follow-up visualization of significant associations

## Related functions

* `add_r_and_stars()`
* `plotSigCorrelations()`
* `MakeComparisonTable()`
* `PlotVolcanoEffects()`
* `CreatePCA()`

## Session information

```{r}
sessionInfo()
```
