---
title: "Threshold-aware p-value color scales"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Threshold-aware p-value color scales}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: VignetteSetup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r}
#| label: LoadPackages
#| message: false

library(SciDataReportR)
library(ggplot2)
library(dplyr)
```

## Motivation

Perceptually uniform palettes such as Viridis and Inferno are excellent
general-purpose continuous color scales. Equal changes in the underlying
quantity produce reasonably consistent visual changes, the palettes remain
legible for many viewers with color-vision deficiencies, and their ordering
survives conversion to greyscale better than many rainbow palettes.

P-values are different from most continuous measurements. Scientists usually
interpret evidence around familiar thresholds such as 0.05, 0.01, and 0.001,
rather than by equal numerical spacing between 0 and 1. A standard linear
color scale gives most of its visual range to large, non-significant p-values
and compresses the region where evidence is commonly compared.

`scale_color_pvalue()` and `scale_fill_pvalue()` therefore use a
threshold-aware mapping. They expect **raw p-values between 0 and 1**, allocate
additional visual resolution near common thresholds, and progressively
desaturate values above 0.05. The legend uses the same warped coordinates, so
its raw p-value labels remain separated around the thresholds instead of
collapsing at one end. Inferno is the default palette.

## Basic usage

Use `scale_color_pvalue()` when a raw p-value is mapped to `color`.

```{r}
#| label: BasicColorScale

df_ColorExample <- data.frame(
  Estimate = seq(-2, 2, length.out = 80),
  PValue = 10^seq(0, -8, length.out = 80)
)

ggplot(
  df_ColorExample,
  aes(x = Estimate, y = 1, color = PValue)
) +
  geom_point(size = 3) +
  scale_color_pvalue() +
  theme_minimal() +
  labs(x = "Estimate", y = NULL)
```

Use `scale_fill_pvalue()` when the p-value controls a filled geometry.

```{r}
#| label: BasicFillScale

df_FillExample <- expand.grid(
  Outcome = paste("Outcome", 1:4),
  Marker = paste("Marker", 1:5)
) %>%
  dplyr::mutate(PValue = 10^seq(0, -6, length.out = dplyr::n()))

ggplot(
  df_FillExample,
  aes(x = Marker, y = Outcome, fill = PValue)
) +
  geom_tile(color = "white") +
  scale_fill_pvalue() +
  theme_minimal() +
  labs(x = NULL, y = NULL)
```

## Volcano plot example

In a volcano plot, the y-axis and color serve different purposes. The y-axis
uses `-log10(PValue)` to place stronger evidence higher on the plot. The color
scale must still receive the raw `PValue`, because the p-value scale performs
its own threshold-aware mapping and displays raw values in the legend.

```{r}
#| label: VolcanoPlot

set.seed(2026)

df_Volcano <- data.frame(
  Feature = paste0("Feature_", seq_len(250)),
  log2FoldChange = stats::rnorm(250),
  PValue = 10^stats::runif(250, min = -7, max = 0)
)

ggplot(
  df_Volcano,
  aes(
    x = log2FoldChange,
    y = -log10(PValue),
    color = PValue
  )
) +
  geom_point(size = 2, alpha = 0.85) +
  scale_color_pvalue() +
  theme_minimal() +
  labs(
    x = expression(log[2] * " fold change"),
    y = expression(-log[10] * "(P-value)")
  )
```

## Bubble plot example

Enrichment plots often use point size for the number of contributing genes
and color for statistical evidence. The color aesthetic again receives the
raw enrichment p-value.

```{r}
#| label: EnrichmentBubblePlot

df_Enrichment <- data.frame(
  Pathway = paste("Pathway", LETTERS[1:10]),
  EnrichmentRatio = seq(1.2, 3.8, length.out = 10),
  GeneCount = c(8, 12, 15, 7, 20, 11, 18, 9, 14, 22),
  PValue = c(0.32, 0.12, 0.049, 0.025, 0.011, 0.006, 0.0018, 0.0007, 1e-4, 1e-6)
)

ggplot(
  df_Enrichment,
  aes(
    x = EnrichmentRatio,
    y = reorder(Pathway, EnrichmentRatio),
    size = GeneCount,
    color = PValue
  )
) +
  geom_point() +
  scale_color_pvalue() +
  theme_minimal() +
  labs(x = "Enrichment ratio", y = NULL, size = "Gene count")
```

## Supported palettes

Inferno is the default. Viridis is available when a cooler palette fits the
surrounding figure better.

```{r}
#| label: SupportedPalettes

ggplot(
  df_ColorExample,
  aes(x = Estimate, y = PValue, color = PValue)
) +
  geom_point(size = 3) +
  scale_color_pvalue(palette = "viridis") +
  scale_y_log10() +
  theme_minimal() +
  labs(y = "P-value")
```

The supported values are:

```r
scale_color_pvalue(palette = "inferno")
scale_color_pvalue(palette = "viridis")
scale_fill_pvalue(palette = "inferno")
scale_fill_pvalue(palette = "viridis")
```

## Advanced options

`limits` sets the raw p-value range represented by the scale. Values outside
the range are squished to the nearest limit. `breaks` and `labels` control the
raw values displayed on the legend. `direction = -1` is the default and makes
smaller p-values darker; `direction = 1` reverses the palette. The default
guide uses a taller vertical colorbar so the threshold labels remain legible.
Pass `guide = "colourbar"` to restore ggplot2's standard guide dimensions, or
provide a custom `ggplot2::guide_colourbar()` for complete control.

```{r}
#| label: AdvancedScaleOptions

ggplot(
  df_Enrichment,
  aes(
    x = EnrichmentRatio,
    y = reorder(Pathway, EnrichmentRatio),
    fill = PValue
  )
) +
  geom_point(shape = 21, size = 6, color = "grey20") +
  scale_fill_pvalue(
    palette = "inferno",
    direction = 1,
    limits = c(1e-6, 1),
    breaks = c(1, 0.05, 0.01, 0.001, 1e-6),
    labels = c("1", "0.05", "0.01", "0.001", "1e-6")
  ) +
  theme_minimal() +
  labs(x = "Enrichment ratio", y = NULL)
```

## Common mistakes

Do not transform the value supplied to the color or fill aesthetic.

```{r}
#| label: IncorrectTransformedColor
#| eval: false

# Incorrect: the p-value scale would receive transformed evidence values.
ggplot(
  df_Volcano,
  aes(
    x = log2FoldChange,
    y = -log10(PValue),
    color = -log10(PValue)
  )
) +
  geom_point() +
  scale_color_pvalue()
```

Instead, transform only the positional axis and pass the raw p-value to the
color scale.

```{r}
#| label: CorrectRawPValueColor
#| eval: false

# Correct: color receives a raw p-value between 0 and 1.
ggplot(
  df_Volcano,
  aes(
    x = log2FoldChange,
    y = -log10(PValue),
    color = PValue
  )
) +
  geom_point() +
  scale_color_pvalue()
```

The same rule applies to `fill`: use `fill = PValue`, not
`fill = -log10(PValue)`. Do not use these scales for effect sizes, fold
changes, correlations, z-scores, probabilities, feature importance, or other
continuous quantities.
