---
title: "Screening predictors with UnivariateRegressionTable()"
description: "Generate publication-ready regression tables and forest plots across multiple outcomes and predictors."
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Screening predictors with UnivariateRegressionTable()}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

`UnivariateRegressionTable()` performs large numbers of regression analyses simultaneously and returns publication-ready summary tables.

This approach is particularly useful for:

* Biomarker screening
* Variable prioritization
* Covariate discovery
* Exploratory analyses
* Hypothesis generation

This vignette demonstrates:

* Running multiple univariate regressions
* Publication-ready tables
* Detailed regression tables
* Covariate adjustment
* Standardized coefficients
* Forest plot visualization

## Load packages

```{r}
library(SciDataReportR)
library(dplyr)
```

## Load example data

```{r}
data("SampleData")
data("SampleVariableTypes")

RevaluedObj <- RevalueData(
  SampleData,
  SampleVariableTypes
)

df_Revalued <- RevaluedObj$RevaluedData
```

## Define outcomes and predictors

For this example we will evaluate a panel of biomarkers as predictors of several clinical outcomes.

```{r}
OutcomeVars <- c(
  "Diagnosis",
  "sex"
)

PredictorVars <- c(
  "Genotype",
  "Calbindin",
  "Ferritin",
  "MMP7",
  "Calbindin",
  "Sortilin",
  "Osteopontin"
)
```

## Create regression tables

```{r}
UniObj <- UnivariateRegressionTable(
  Data = df_Revalued,
  OutcomeVars = OutcomeVars,
  PredictorVars = PredictorVars
)
```

## Publication-ready table

The formatted table is designed for reporting and manuscript preparation.

```{r}
UniObj$FormattedTable
```

Estimates and standard errors are combined into a compact format and significance stars are automatically added.

## Detailed regression table

The larger table provides additional model details.

```{r}
UniObj$LargeTable
```

This version is often useful during exploratory analyses and quality control.

## Including covariates

Covariates can be included in every regression model.

For example, age is commonly included when evaluating biomarker associations.

```{r}
UniObj_Covar <- UnivariateRegressionTable(
  Data = df_Revalued,
  OutcomeVars = OutcomeVars,
  PredictorVars = PredictorVars,
  Covars = "age"
)
```

```{r}
UniObj_Covar$FormattedTable
```

This allows investigators to determine whether associations remain significant after accounting for potential confounding factors.

## Standardized coefficients

When predictors are measured on different scales, standardized coefficients can simplify interpretation.

```{r}
UniObj_Std <- UnivariateRegressionTable(
  Data = df_Revalued,
  OutcomeVars = OutcomeVars,
  PredictorVars = PredictorVars,
  Standardize = TRUE
)
```

```{r}
UniObj_Std$FormattedTable
```

Standardized coefficients represent changes in standard deviation units and facilitate comparison across predictors.

## Accessing model objects

The fitted regression models are returned in the output object.

```{r}
names(
  UniObj$ModelSummaries
)
```

For example:

```{r}
summary(
  UniObj$ModelSummaries$MMSE$AXL
)
```

This allows additional diagnostics and custom analyses.

## Creating a forest plot

A forest plot provides a compact visual summary of regression results.

```{r}
ForestPlot <- plotForestFromTable(
  UniObj
)

ForestPlot
```

Forest plots make it easy to identify:

* Strong predictors
* Significant predictors
* Direction of effects
* Precision of estimates

## Forest plot with covariate adjustment

The same visualization can be generated using adjusted models.

```{r}
plotForestFromTable(
  UniObj_Covar
)
```

Comparing adjusted and unadjusted models can help identify associations that may be explained by confounding variables.

## Recommended workflow

A common workflow is:

1. Use `UnivariateRegressionTable()` to screen large numbers of predictors.
2. Review the formatted table.
3. Compare standardized and unstandardized results.
4. Add important covariates.
5. Visualize findings using `plotForestFromTable()`.
6. Follow up significant findings using multivariable models.

## Summary

`UnivariateRegressionTable()` provides a rapid screening framework for evaluating many predictors across multiple outcomes.

Key features include:

* Multiple outcomes
* Multiple predictors
* Optional covariate adjustment
* Standardized coefficients
* Publication-ready tables
* Detailed regression output
* Forest plot visualization

## Related functions

* `plotForestFromTable()`
* `MakeComparisonTable()`
* `PlotMiningMatrix()`
* `PlotCorrelationsHeatmap()`
* `CreatePCA()`

## Session information

```{r}
sessionInfo()
```
