---
title: "Merge QC workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Merge QC workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
has_survival <- requireNamespace("survival", quietly = TRUE)
has_widgets <- requireNamespace("reactable", quietly = TRUE) &&
  requireNamespace("htmltools", quietly = TRUE)
knitr::opts_chunk$set(eval = has_survival)
```

Merging datasets is where many analysis errors are born: silently duplicated
rows, IDs that never match, and `.x` / `.y` column pairs that nobody resolves.
SciDataReportR wraps the merge-and-audit cycle into a small workflow:

- `safe_merge()` performs a left join (exact-key or closest-time) and
  immediately audits the result with `ValidateMerge()`.
- `merge_detail()` prints plain-text diagnostic tables for one merge.
- `ExploreMergeValidation()` renders an interactive dashboard for one merge.
- `merge_summary_table()` stacks the one-row logs from many merges into a
  pipeline-level rollup.

```{r setup, message = FALSE}
library(SciDataReportR)
library(survival)
```

## Example 1: a real longitudinal merge (survival::pbc + pbcseq)

The `survival` package ships two related datasets from the Mayo Clinic primary
biliary cholangitis trial:

- `pbc`: baseline data, one row per patient (418 patients, including 106 who
  did not participate in the randomized trial),
- `pbcseq`: longitudinal follow-up labs, multiple rows per patient (only the
  312 trial participants).

Merging them by patient `id` is a completely ordinary task — and it surfaces
real coverage gaps and duplicate-key situations that were not staged for this
vignette.

```{r}
pbc_baseline <- pbc[, c("id", "age", "sex", "trt", "stage")]
pbc_labs <- pbcseq[, c("id", "day", "bili", "albumin", "platelet")]

m_pbc <- safe_merge(
  pbc_baseline,
  pbc_labs,
  by = "id",
  name = "pbc baseline + follow-up labs"
)
```

`safe_merge()` returns four elements and prints nothing on its own. The
one-row `log` is the pipeline-friendly view:

```{r}
m_pbc$log
```

The `summary` element is a ready-made `knitr::kable()`:

```{r}
m_pbc$summary
```

The audit correctly reports this merge as `FAIL` — not because the code is
wrong, but because the result needs a decision from you before analysis:

- **Coverage**: 106 patients exist only in the baseline data (the
  non-trial patients who have no follow-up labs).
- **Duplicate keys**: `pbcseq` has many rows per patient, so `id` alone does
  not uniquely identify a row after the merge. If you intended a one-row-per-
  patient dataset, this is a genuine error; if you intended a longitudinal
  dataset, you now have written confirmation of what happened.

### Drilling in with merge_detail()

`merge_detail()` prints plain kable tables for the checks, the unmatched keys
on each side, overlapping variables, and suspicious conflicts, skipping any
empty section. In R Markdown, use a chunk with `results = "asis"`:

```{r, results = "asis"}
merge_detail(m_pbc, TopN = 5)
```

### A closest-time variant

Suppose instead you want a one-row-per-patient dataset containing the lab
draw closest to enrollment. `safe_merge(method = "closest_time")` calls
`Merge_ByClosestTime()` under the hood, matching on `id` exactly and picking
the `pbc_labs` row whose `day` is nearest to each baseline row's time
(enrollment is day 0 here).

```{r}
pbc_enroll <- pbc_baseline
pbc_enroll$enroll_day <- 0

m_closest <- safe_merge(
  pbc_enroll,
  pbc_labs,
  by = "id",
  name = "closest lab to enrollment",
  method = "closest_time",
  time_var_before = "enroll_day",
  time_var_add = "day"
)

m_closest$log
```

Note the caveat: `ValidateMerge()` audits duplicate keys using `by` alone
(here `id`), not the time variables. In longitudinal settings, repeated keys
in the lab data are legitimate repeated visits, so a duplicate-key `FAIL` from
a closest-time merge may reflect expected repetition rather than a merge
error. The validation output is deliberately left as-is — read it with that
caveat in mind (here, the merged data itself has one row per patient, which
you can confirm from `RowsBefore` and `RowsAfter`).

## Example 2: a synthetic merge that fails every check

To demonstrate the remaining check types — key-type mismatches, unresolved
`.x` / `.y` pairs, and value conflicts — here is a small synthetic pair of
data frames constructed to go wrong in all the classic ways: a duplicated
key, an ID stored as integer on one side and double on the other, IDs
missing from each side, and a `sex` column present in both sources with
disagreeing values.

```{r}
demographics <- data.frame(
  id = 1:6, # integer
  sex = c("F", "M", "F", "F", "M", "F"),
  age = c(54, 61, 47, 66, 58, 50)
)

device_data <- data.frame(
  id = c(1, 2, 2, 5, 7), # double, with id 2 duplicated
  sex = c("F", "F", "M", "M", "F"), # disagrees with demographics for id 2/5
  score = c(0.82, 0.75, 0.71, 0.64, 0.90)
)

m_synth <- safe_merge(
  demographics,
  device_data,
  by = "id",
  name = "demographics + device data"
)

m_synth$log
```

```{r}
m_synth$summary
```

Every check type in `ValidateMerge()` now has something to say: the duplicate
key and the unresolved `sex.x` / `sex.y` pair are integrity blockers
(`FAIL`), while key-type coercion, coverage gaps, row inflation, overlapping
variables, and value conflicts appear as warnings.

```{r, results = "asis"}
merge_detail(m_synth)
```

### Interactive review with ExploreMergeValidation()

For interactive QC sessions, pass the `validation` element to
`ExploreMergeValidation()`. The default `Detail = "Compact"` keeps the checks
table front and center, with the coverage and conflict explorers as
click-to-expand accordion sections labeled with their item counts. Use
`Detail = "Full"` to render them expanded.

```{r, eval = has_survival && has_widgets}
ExploreMergeValidation(
  m_synth$validation,
  Title = "Demographics + device data",
  Detail = "Compact"
)
```

## Pipeline rollup with merge_summary_table()

After a sequence of merges, stack the logs into a single table. With
`flagged_only = TRUE`, only merges whose worst check status is not `PASS`
remain — an end-of-pipeline QC gate:

```{r}
merge_summary_table(
  list(m_pbc$log, m_closest$log, m_synth$log),
  flagged_only = TRUE
)
```

An empty table here would mean every merge in the pipeline passed cleanly;
any row that appears is a merge that still needs review before analysis.
